---
title: "Take-Home Exercise 01"
author: "BingBing"
date: "2025-09-20"
format: html
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1. Overview

This exercise applies spatial point pattern analysis methods to study the distribution of newly registered corporate entities in Singapore. Following the Data Analytics Lifecycle and based on the hands-on practices, the analysis covers data preparation, industry classification, geocoding, and geovisualisation.

# 2. **Installing and Launching the R Packages**

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|-------------|-------------|--------------------------|

\
\
\
\
\

```{r}
pacman::p_load(tidyverse, readr, lubridate,sf,tmap, rvest,httr, jsonlite, dplyr, purrr,spatstat)
```

## 2.1 **Reproducibilty**

For reproducible results of this exercise, we will use seed value, `1234`.

```{r}
set.seed(1234)
```

# 3. Data Import & Preparation

The following 2 datasets will be used in this exercise.

| **Dataset Name** | **Description** | **Format** | **Source** |
|------------------|-----------------|------------|------------|

-   ACRA datasets

-   postal code\

Import ACRA datasets from **data.gov.sg**.

```{r}
acra_files <- list.files("data/aspatial/ACRA", 
                         pattern = "*.csv", 
                         full.names = TRUE)
```

```{r}
acra_raw <- map_dfr(acra_files, read_csv)
```

Select relevant fields and filter entities registered between **1 Jan 2024 and 30 Jun 2025**.

```{r}
acra_clean <- acra_raw %>%
  select(
    uen,
    entity_name,
    ssic_code = primary_ssic_code,
    ssic_desc = primary_ssic_description,
    postal = postal_code,
    reg_date = registration_incorporation_date
  ) %>%
  filter(reg_date >= "2024-01-01", reg_date <= "2025-06-30") %>%
  filter(!is.na(postal)) %>%
  distinct(uen, .keep_all = TRUE)   
   
glimpse(acra_clean)
```

According to SSIC 2020 (Singapore Standard Industrial Classification), this study selected three industry categories:

-   56 (Food and Beverage Service Activities)

-   62 (Computer Programming and IT Consultancy Activities)

-   85 (Education)

```{r}
# Select the industry according to the official classification of SSIC 2020
chosen_ssic <- c("56", "62", "85")

industry_labels <- c(
  "56" = "Food & Beverage",
  "62" = "IT Services",
  "85" = "Education")

acra_selected <- acra_clean %>%
  mutate(ssic_2d = str_sub(ssic_code, 1, 2)) %>%
  filter(ssic_2d %in% chosen_ssic)

acra_selected %>%
  count(ssic_2d)

```

```{#rds file}
if(!dir.exists("data/aspatial/kde_rds")) {
  dir.create("data/aspatial/kde_rds", recursive = TRUE)
}

for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    
    bw <- bw.diggle(ppp_ind)  
    kde_ind <- density.ppp(ppp_ind, sigma = bw)  # KDE
    saveRDS(kde_ind, paste0("data/aspatial/kde_rds/kde_", ind, ".rds"))
  }
}
```

```{industries <- c("56", "62", "85")}
coords_all <- st_coordinates(acra_sf_proj)
ppp_all <- spatstat.geom::ppp(
  x = coords_all[,1],
  y = coords_all[,2],
  window = spatstat.geom::as.owin(mp19_subzone_proj)
)
bw_global <- bw.diggle(ppp_all)
for (ind in industries) {
  sub_ind <- acra_sf_proj %>% filter(ssic_2d == ind)
  sub_ind <- sub_ind %>%
    mutate(year_qtr = paste0(year(reg_date), "Q", quarter(reg_date)))
  
  for (q in unique(sub_ind$year_qtr)) {
    sub_q <- sub_ind %>% filter(year_qtr == q)
    coords <- st_coordinates(sub_q)
    
    if (nrow(coords) > 30) {
      ppp_q <- spatstat.geom::ppp(
        x = coords[,1], 
        y = coords[,2],
        window = spatstat.geom::as.owin(mp19_subzone_proj)
      )
      
      kde_q <- density.ppp(ppp_q, sigma = bw_global)
      saveRDS(kde_q, file = paste0("data/aspatial/kde_rds/kde_", ind, "_", q, ".rds"))
    }
  }
}

```

Load **Master Plan 2019 Subzone Boundary (No Sea)** for spatial reference.

```{r}
mp19_subzone <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
mp19_subzone <- st_zm(mp19_subzone, drop = TRUE, what = "ZM")
```

check the data:

```{r}
glimpse(mp19_subzone)
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders()
```

Use the OneMap API (2019) to geocode postal codes. Please register on the OneMap portal to obtain your own API token. Geocoded results were saved as `data/aspatial/acra_postal_coords.csv`. Merge geocoded coordinates and convert to sf object.

```{r}
coords_all <- read_csv("data/aspatial/acra_postal_coords.csv")

acra_geo <- acra_selected %>%
  left_join(coords_all, by = "postal") %>%
  drop_na(lat, lon)

acra_sf <- st_as_sf(acra_geo, coords = c("lon", "lat"), crs = 4326)
acra_sf <- st_transform(acra_sf, st_crs(mp19_subzone))
```

you can check the output by :

```{r}
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders() +
  tm_shape(acra_sf) + tm_dots(col = "ssic_2d", size = 0.05)
```

# 4. Spatial and Spatio-temporal KDE

Kernel Density Estimation (KDE) is applied to identify the spatial concentration of newly registered corporate entities. This section follows the approach in Hands-on Exercise.
```{r}
acra_sf_proj <- st_transform(acra_sf, 3414)
mp19_subzone_proj <- st_transform(mp19_subzone, 3414)
```

## 4.1 Spatial KDE by Industry

```{r}
#| fig.height: 9
#| fig.width: 9
#| warning: false

industries <- c("56", "62", "85")
for (ind in industries) {
  rds_file <- paste0("data/aspatial/kde_rds/kde_", ind, ".rds")
  
  if (file.exists(rds_file)) {
    kde_ind <- readRDS(rds_file)
    
    plot(kde_ind, main = paste("KDE -", industry_labels[ind]))
    plot(mp19_subzone_proj$geometry, add = TRUE)
  }
}

```

## 4.2 Spatio-temporal KDE by Industry (Quarterly)

```{r}
#| fig.height: 9
#| fig.width: 9
#| warning: false

for (ind in industries) {
  files <- list.files("data/aspatial/kde_rds", pattern = paste0("kde_", ind, "_"), full.names = TRUE)
  
  for (f in files) {
    kde_q <- readRDS(f)
    q <- str_extract(f, "20\\d{2}Q[1-4]")
    
    plot(kde_q, main = paste("KDE -", industry_labels[ind], q))
    plot(mp19_subzone_proj$geometry, add = TRUE)
  }
}

```

#### 

## 5. Second-order Analysis (Per Industry & Cross-industry)

The quadrat test is used to statistically evaluate whether the spatial distribution of business entities within each industry follows Complete Spatial Randomness (CSR). A significant p-value (e.g., \< 0.05) indicates deviation from CSR, suggesting clustering or regularity.

## 5.1 Quadrat Test by Industry

```{r}
#| fig.height: 9
#| fig.width: 9
#| warning: false

for (ind in industries) {
  rds_file <- paste0("data/aspatial/kde_rds/quadrat_", ind, ".rds")
  
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], 
      y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
 
    if (!file.exists(rds_file)) {
      qtest_ind <- quadrat.test(ppp_ind)
      saveRDS(qtest_ind, rds_file)
    }
    
    qtest_ind <- readRDS(rds_file)
    
    print(paste("Quadrat test result for industry:", industry_labels[ind]))
    print(qtest_ind)
    plot(ppp_ind, main = paste("Quadrat Test -", industry_labels[ind]))
    plot(quadratcount(ppp_ind, nx = 4, ny = 4), add = TRUE)
  }
}

```

### 5.2 Nearest Neighbour Analysis

The nearest neighbour functions provide second-order summaries of spatial point patterns:
- **G-function**: Distribution of nearest neighbour distances.  
- **F-function**: Empty space distribution.  
- **K-function**: Measures clustering or dispersion at different scales.  

We compare the observed functions with the theoretical expectation under CSR (Complete Spatial Randomness).
```{r}
#| warning: false


for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], 
      y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    
    # 计算 G, F, K
    g_res <- Gest(ppp_ind)
    f_res <- Fest(ppp_ind)
    k_res <- Kest(ppp_ind)
    
  
    par(mfrow=c(1,3))  
    plot(g_res, main = paste("G-function -", industry_labels[ind]))
    plot(f_res, main = paste("F-function -", industry_labels[ind]))
    plot(k_res, main = paste("K-function -", industry_labels[ind]))
    par(mfrow=c(1,1)) 
  }
}

```
```{r}
#| warning: false

for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    
    g_res <- Gest(ppp_ind)
    plot(g_res, main = paste("G-function -", industry_labels[ind]))
  }
}

```
```{r}
#| warning: false

for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    
    f_res <- Fest(ppp_ind)
    plot(f_res, main = paste("F-function -", industry_labels[ind]))
  }
}
```

```{r}
#| warning: false

for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    
    k_res <- Kest(ppp_ind)
    plot(k_res, main = paste("K-function -", industry_labels[ind]))
  }
}
```


```{r}

industries <- c("56", "62", "85")


if(!dir.exists("data/aspatial/kde_rds")) {
  dir.create("data/aspatial/kde_rds", recursive = TRUE)
}

for (ind in industries) {
  sub <- acra_sf_proj %>% filter(ssic_2d == ind)
  coords <- st_coordinates(sub)
  
  if (nrow(coords) > 30) {
    ppp_ind <- spatstat.geom::ppp(
      x = coords[,1], 
      y = coords[,2],
      window = spatstat.geom::as.owin(mp19_subzone_proj)
    )
    

    rds_file <- paste0("data/aspatial/kde_rds/Lenv_", ind, ".rds")
    
    if (file.exists(rds_file)) {
      L_env <- readRDS(rds_file)
    } else {
      L_env <- envelope(ppp_ind, Lest, nsim = 99, rank = 1, global = TRUE,
                        savepatterns = FALSE, verbose = FALSE)
      saveRDS(L_env, rds_file)
    }
    
    plot(L_env, . - r ~ r,
         main = paste("L-function (CSR Envelope) -", industry_labels[ind]),
         xlab = "d", ylab = "L(d) - r")
  }
}


```




##### 5.3 Cross-K (可选)

## 5.1 G/F/K/L Functions per Industry


5.2 Monte Carlo Envelopes (CSR Test)


5.3 Cross-industry Interaction (Kcross / Gcross)

