---
title: "Take-Home Exercise 01"
author: "BingBing"
date: "2025-09-20"
format: html
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

# 1. Overview

This exercise applies spatial point pattern analysis methods to study the distribution of newly registered corporate entities in Singapore. Following the Data Analytics Lifecycle and based on the hands-on practices, the analysis covers data preparation, industry classification, geocoding, and geovisualisation.

# 2. **Installing and Launching the R Packages**

The following R packages will be used in this exercise:

| **Package** | **Purpose** | **Use Case in Exercise** |
|-------------|-------------|--------------------------|

\
\
\
\
\

```{r}
pacman::p_load(tidyverse, readr, lubridate,sf,tmap, rvest,httr, jsonlite, dplyr, purrr,spatstat)
```

## 2.1 **Reproducibilty**

For reproducible results of this exercise, we will use seed value, `1234`.

```{r}
set.seed(1234)
```

# 3. Data Import & Preparation

The following 2 datasets will be used in this exercise.

| **Dataset Name** | **Description** | **Format** | **Source** |
|------------------|-----------------|------------|------------|

-   ACRA (Accounting and Corporate Regulatory Authority) Information on Corporate Entities datasets from data.gov.sg

-   postal code

-   Master Plan 2019 Subzone Boundary (No Sea), a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

Import ACRA datasets from **data.gov.sg**.

```{r}
acra_files <- list.files("data/aspatial/ACRA", 
                         pattern = "*.csv", 
                         full.names = TRUE)
```

```{r}
acra_raw <- map_dfr(acra_files, read_csv)
```

Select relevant fields and filter entities registered between **1 Jan 2024 and 30 Jun 2025**.

```{r}
acra_clean <- acra_raw %>%
  select(
    uen,
    entity_name,
    ssic_code = primary_ssic_code,
    ssic_desc = primary_ssic_description,
    postal = postal_code,
    reg_date = registration_incorporation_date
  ) %>%
  filter(reg_date >= "2024-01-01", reg_date <= "2025-06-30") %>%
  filter(!is.na(postal)) %>%
  distinct(uen, .keep_all = TRUE)   
   
glimpse(acra_clean)
```

According to SSIC 2020 (Singapore Standard Industrial Classification), this study selected three industry categories:

-   56 (Food and Beverage Service Activities)

-   62 (Computer Programming and IT Consultancy Activities)

-   85 (Education)

```{r}
# Select the industry according to the official classification of SSIC 2020
chosen_ssic <- c("56", "62", "85")

industry_labels <- c(
  "56" = "Food & Beverage",
  "62" = "IT Services",
  "85" = "Education")

acra_selected <- acra_clean %>%
  mutate(ssic_2d = str_sub(ssic_code, 1, 2)) %>%
  filter(ssic_2d %in% chosen_ssic)

acra_selected %>%
  count(ssic_2d)

```


Load **Master Plan 2019 Subzone Boundary (No Sea)** for spatial reference.

```{r}
mp19_subzone <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
mp19_subzone <- st_zm(mp19_subzone, drop = TRUE, what = "ZM")
```

check the data:

```{r}
glimpse(mp19_subzone)
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders()
```

Use the OneMap API (2019) to geocode postal codes. Please register on the OneMap portal to obtain your own API token. Geocoded results were saved as `data/aspatial/acra_postal_coords.csv`. Merge geocoded coordinates and convert to sf object.

```{r}
coords_all <- read_csv("data/aspatial/acra_postal_coords.csv")

acra_geo <- acra_selected %>%
  left_join(coords_all, by = "postal") %>%
  drop_na(lat, lon)

acra_sf <- st_as_sf(acra_geo, coords = c("lon", "lat"), crs = 4326)
acra_sf <- st_transform(acra_sf, st_crs(mp19_subzone))
```

you can check the output by :

```{r}
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders() +
  tm_shape(acra_sf) + tm_dots(col = "ssic_2d", size = 0.05)
```

```{r}
acra_sf_proj <- st_transform(acra_sf, 3414)
mp19_subzone_proj <- st_transform(mp19_subzone, 3414)
```

 
4. Geospatial Data wrangling

spatstat relies on its own specific data structures like ppp (planar point pattern) for point data and owin for observation windows


4.1 Converting sf data frames to ppp class
spatstat requires the point event data in ppp object form. The code chunk below uses [as.ppp()] of spatstat package to convert childcare_sf to ppp format.
```{r}
acra_ppp <- as.ppp(acra_sf_proj)
class(acra_ppp)
```



```{r}
summary(acra_ppp)
```


4.2 Creating owin object
When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

The code chunk below, as.owin() of spatstat is used to covert mpsz_sf into owin object of spatstat.

```{r}
sg_owin <- as.owin(mp19_subzone_proj)
class(sg_owin)
plot(sg_owin)
```
4.2.1 Combining point events object and owin object
```{r}
sg_owin <- as.owin(mpsz_cl)
```


```{r}
acra_ppp = childcare_ppp[sg_owin]
```







