---
title: "Take-Home Exercise 01"
author: "BingBing"
date: "2025-09-20"
format: html
execute: 
  eval: true
  echo: false
  warning: false
  freeze: true
---

# 1. Overview

This exercise applies spatial point pattern analysis methods to study the distribution of newly registered corporate entities in Singapore. Following the Data Analytics Lifecycle and based on the hands-on practices, the analysis covers data preparation, industry classification, geocoding, and geovisualisation.

# 2. **Installing and Launching the R Packages**

```{r}
pacman::p_load(tidyverse, readr, lubridate,sf,tmap, rvest,httr, jsonlite, dplyr, purrr,spatstat,terra,spatstat.explore,stars,spatstat.geom,rlang)

```

## 2.1 **Reproducibilty**

For reproducible results of this exercise, we will use seed value, `1234`.

```{r}
set.seed(1234)
```

# 3. Data Import & Preparation

The following 3 datasets will be used in this exercise.

-   ACRA (Accounting and Corporate Regulatory Authority) Information on Corporate Entities datasets from data.gov.sg

-   postal code download by onemap

-   Master Plan 2019 Subzone Boundary (No Sea), a polygon feature data providing information of URA 2019 Master Plan Planning Subzone boundary data.

Import ACRA datasets from **data.gov.sg**.

```{r}
acra_files <- list.files("data/aspatial/ACRA", 
                         pattern = "*.csv", 
                         full.names = TRUE)
```

```{r}

acra_raw <- map_dfr(acra_files, read_csv)
```

Select relevant fields and filter entities registered between **1 Jan 2024 and 30 Jun 2025**.

```{r}
acra_clean <- acra_raw %>%
  select(
    uen,
    entity_name,
    ssic_code = primary_ssic_code,
    ssic_desc = primary_ssic_description,
    postal = postal_code,
    reg_date = registration_incorporation_date
  ) %>%
  filter(reg_date >= "2024-01-01", reg_date <= "2025-06-30") %>%
  filter(!is.na(postal)) %>%
  distinct(uen, .keep_all = TRUE)   
   
glimpse(acra_clean)
```

According to SSIC 2020 (Singapore Standard Industrial Classification), this study selected three industry categories:

-   56 (Food and Beverage Service Activities)

-   62 (Computer Programming and IT Consultancy Activities)

-   85 (Education)

```{r}
# Select the industry according to the official classification of SSIC 2020
chosen_ssic <- c("56", "62", "85")

industry_labels <- c(
  "56" = "Food & Beverage",
  "62" = "IT Services",
  "85" = "Education")

acra_selected <- acra_clean %>%
  mutate(ssic_2d = str_sub(ssic_code, 1, 2)) %>%
  filter(ssic_2d %in% chosen_ssic)

acra_selected %>%
  count(ssic_2d)

```

Load **Master Plan 2019 Subzone Boundary (No Sea)** for spatial reference.

```{r}
mp19_subzone <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml")
mp19_subzone <- st_zm(mp19_subzone, drop = TRUE, what = "ZM")
```

check the data:

```{r}
glimpse(mp19_subzone)
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders()
```

Use the OneMap API (2019) to geocode postal codes. Please register on the OneMap portal to obtain your own API token. Geocoded results were saved as `data/aspatial/acra_postal_coords.csv`. Merge geocoded coordinates and convert to sf object.

```{r}
coords_all <- read_csv("data/aspatial/acra_postal_coords.csv")

acra_geo <- acra_selected %>%
  left_join(coords_all, by = "postal") %>%
  drop_na(lat, lon)

acra_sf <- st_as_sf(acra_geo, coords = c("lon", "lat"), crs = 4326)
acra_sf <- st_transform(acra_sf, st_crs(mp19_subzone))
```

you can check the output by :

```{r}
tmap_mode("view")
tm_shape(mp19_subzone) + tm_borders() +
  tm_shape(acra_sf) + tm_dots(col = "ssic_2d", size = 0.05)
```

```{r}
acra_sf_proj <- st_transform(acra_sf, 3414)
mp19_subzone_proj <- st_transform(mp19_subzone, 3414)
```

# 4.Geospatial Data wrangling

spatstat relies on its own specific data structures like ppp (planar point pattern) for point data and owin for observation windows

## 4.1 Converting sf data frames to ppp class

spatstat requires the point event data in ppp object form. The code chunk below uses \[as.ppp()\] of spatstat package to convert childcare_sf to ppp format.

```{r}
acra_ppp <- as.ppp(acra_sf_proj)
class(acra_ppp)
```

```{r}
summary(acra_ppp)
```

## 4.2 Creating owin object

When analysing spatial point patterns, it is a good practice to confine the analysis with a geographical area like Singapore boundary. In spatstat, an object called owin is specially designed to represent this polygonal region.

```{r}
sg_owin <- as.owin(mp19_subzone_proj)
class(sg_owin)
plot(sg_owin)
```

### 4.2.1 Combining point events object and owin object

```{r}
sg_owin <- as.owin(mp19_subzone_proj)
```

```{r}
# Education
acra_85 <- acra_sf_proj %>% filter(ssic_2d == "85")
acra_85_ppp <- as.ppp(acra_85)
acra_85SG_ppp <- acra_85_ppp[sg_owin]

acra_85SG_ppp
```

```{r}
# Food & Beverage
acra_56 <- acra_sf_proj %>% filter(ssic_2d == "56")
acra_56_ppp <- as.ppp(acra_56)
acra_56SG_ppp <- acra_56_ppp[sg_owin]

acra_56SG_ppp
```

```{r}
# IT Services
acra_62 <- acra_sf_proj %>% filter(ssic_2d == "62")
acra_62_ppp <- as.ppp(acra_62)
acra_62SG_ppp <- acra_62_ppp[sg_owin]

acra_62SG_ppp
```

## 4.3 Clark-Evan Test for Nearest Neighbour Analysis

The Clark-Evans test (Nearest Neighbour Analysis, NNA) is a spatial statistics method that calculates the average distance between each point and its closest neighbor to determine if a pattern of points is clustered, dispersed, or randomly distributed.\

We will perform the Clark-Evans test of aggregation for each selected industry by using `clarkevans.test()` of **spatstat.explore** package.

The general test hypotheses are:

-   **H0**: The distribution of business entities is randomly distributed (CSR – Complete Spatial Randomness).

-   **H1**: The distribution of business entities is not randomly distributed (i.e., significantly clustered or dispersed).

The 95% confidence interval will be used.

### **4.3.1 Perform the Clark-Evans test without CSR**

#### 4.3.1.1 Food & Beverage (SSIC 56)

For **Food & Beverage** industry, the hypotheses are:

-   **H0**: The distribution of Food & Beverage businesses is randomly distributed.

-   **H1**: The distribution of Food & Beverage businesses is not randomly distributed.

```{r}
#| fig.height: 6
#| fig.width: 6
#| warning: false
#| eval: false

sub_fnb <- acra_sf_proj %>% filter(ssic_2d == "56")
coords_fnb <- st_coordinates(sub_fnb)
ppp_fnb <- spatstat.geom::ppp(x = coords_fnb[,1], y = coords_fnb[,2],window = sg_owin)

clarkevans.test(ppp_fnb,
                correction = "none",
                clregion = "sg_owin",
                alternative = c("clustered"))
```

**Insights:**

-   Clark-Evans test gives **R = 0.36, p \< 0.001**, shows **significantly clustered**.

-   This means F&B outlets are not randomly distributed but significantly concentrated in certain locations.

#### 4.3.1.2 IT Services (SSIC 62)

For **IT Services** industry, the hypotheses are:

-   **H0**: The distribution of IT Services businesses is randomly distributed.

-   **H1**: The distribution of IT Services businesses is not randomly distributed.

```{r}
#| fig.height: 6
#| fig.width: 6
#| warning: false
#| eval: false

sub_it <- acra_sf_proj %>% filter(ssic_2d == "62")
coords_it <- st_coordinates(sub_it)
ppp_it <- spatstat.geom::ppp(x = coords_it[,1], y = coords_it[,2],window = sg_owin)

clarkevans.test(ppp_it,
                correction = "none",
                clregion = "sg_owin",
                alternative = c("clustered"))
```

**Insights:**

-   Clark-Evans test gives **R = 0.25, p \< 0.001**, indicating **strong clustering**.

-   The distribution is highly concentrated, suggesting strong spatial dependence.

#### 4.3.1.3 Education (SSIC 85)

For **Education** industry, the hypotheses are:

H0: The distribution of Education businesses is randomly distributed.

H1: The distribution of Education businesses is not randomly distributed.

```{r}
#| fig.height: 6
#| fig.width: 6
#| warning: false
#| eval: false

sub_edu <- acra_sf_proj %>% filter(ssic_2d == "85")
coords_edu <- st_coordinates(sub_edu)
ppp_edu <- spatstat.geom::ppp(x = coords_edu[,1], y = coords_edu[,2],window = sg_owin)

clarkevans.test(ppp_edu,
                correction = "none",
                clregion = "sg_owin",
                alternative = c("clustered"))

```

**Insights:**

-   Clark-Evans test gives **R = 0.37, p \< 0.001**, also **significantly clustered**.

-   Education providers tend to be located in specific zones rather than evenly across Singapore.

### **4.3.2 Perform the Clark-Evans test with CSR**

In the code chunk below, the argument method = “MonteCarlo” is used. In this case, the p-value for the test is computed by comparing the observed value of R to the results obtained from nsim (i.e. 39, 99, 999) simulated realisations of Complete Spatial Randomness conditional on the observed number of points.

```{r}
#| warning: false
#| eval: false

# Food & Beverage
ce_fnb <- clarkevans.test(ppp_fnb,
                          correction = "none",
                          clipregion = sg_owin,
                          alternative = c("clustered"),
                          method = "MonteCarlo",
                          nsim = 99)
saveRDS(ce_fnb, "data/aspatial/kde_rds/CE_FNB.rds")

# IT Services
ce_it <- clarkevans.test(ppp_it,
                         correction = "none",
                         clipregion = sg_owin,
                         alternative = c("clustered"),
                         method = "MonteCarlo",
                         nsim = 99)
saveRDS(ce_it, "data/aspatial/kde_rds/CE_IT.rds")

# Education
ce_edu <- clarkevans.test(ppp_edu,
                          correction = "none",
                          clipregion = sg_owin,
                          alternative = c("clustered"),
                          method = "MonteCarlo",
                          nsim = 99)
saveRDS(ce_edu, "data/aspatial/kde_rds/CE_EDU.rds")

```

```{r}
ce_fnb <- readRDS("data/aspatial/kde_rds/CE_FNB.rds")
ce_it  <- readRDS("data/aspatial/kde_rds/CE_IT.rds")
ce_edu <- readRDS("data/aspatial/kde_rds/CE_EDU.rds")
```

```{r}
ce_fnb
```

-   The significant R \< 1 indicates strong clustering. F&B businesses tend to group around popular dining districts and residential catchments, reflecting demand-driven location strategies.

```{r}
ce_it
```

-   The lowest R among the three industries, suggesting the strongest clustering. This likely reflects concentration in CBDs and business parks where digital and consultancy firms prefer proximity to clients and talent.

```{r}
ce_edu
```

-   Also shows clear clustering, though slightly weaker than IT. This may align with population density, as educational services are concentrated in residential towns and central areas.

## **5. Kernel Density Estimation Method**

**Kernel Density Estimation (KDE)** is a valuable tool for visualising and analyzing first-order spatial point patterns. It is widely considered a method within **Exploratory Spatial Data Analysis (ESDA).** We will use Kernel Density Estimation (KDE) is applied to visualize the spatial intensity of newly registered businesses. It helps us **identify hotspots, highlight areas with high clustering**, and provides a clearer picture of the spatial distribution pattern beyond point maps.

**5.1 Working with automatic bandwidth selection method**

Bandwidth selectors：

-   bw.diggle(): conservative (controls false positives); good general default.

-   bw.CvL(): Cressie–Read cross-validation; tends to optimize predictive fit.

-   bw.scott(): rule-of-thumb; reasonable for near-uniform patterns.

-   bw.ppl(): profile pseudo-likelihood; often a robust first choice.

-   Kernel: use gaussian (default); alternatives: "epanechnikov", "quartic", "disc".

-   Edge correction: set edge = TRUE (Jones, 1993; Diggle, 2010) to reduce boundary underestimation.

-   Coordinate system: run KDE in a metric projection (e.g., EPSG:3414) because it depends on distances.

-   Units: density.ppp() returns intensity per m²; multiply by 1e6 (10\^6) to report per km².

```{r}
# Food & Beverage
kde_SG_fnb <- density(
  acra_56SG_ppp,
  sigma = bw.diggle(acra_56SG_ppp),
  edge = TRUE,
  kernel = "gaussian"
)

# IT Services
kde_SG_it <- density(
  acra_62SG_ppp,
  sigma = bw.diggle(acra_62SG_ppp),
  edge = TRUE,
  kernel = "gaussian"
)

# Education
kde_SG_edu <- density(
  acra_85SG_ppp,
  sigma = bw.diggle(acra_85SG_ppp),
  edge = TRUE,
  kernel = "gaussian"
)

```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
kde_SG_fnb <- density(
  acra_56SG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 

plot(kde_SG_fnb)
```

```{r}
summary(kde_SG_fnb)
```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

kde_SG_it <- density(
  acra_62SG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 

plot(kde_SG_it)

```

```{r}
summary(kde_SG_it)
```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
kde_SG_edu <- density(
  acra_85SG_ppp,
  sigma=bw.diggle,
  edge=TRUE,
  kernel="gaussian") 

plot(kde_SG_edu)
```

```{r}
summary(kde_SG_edu)
```

**5.2 Rescalling KDE values**

In the code chunk below, *rescale.ppp()* is used to covert the unit of measurement from meter to kilometer.

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
# Food & Beverage
ppp_fnb <- acra_56SG_ppp
ppp_fnb_km <- rescale.ppp(ppp_fnb, 1000, "km")
kde_fnb <- density(
  ppp_fnb_km,
  sigma = bw.diggle(ppp_fnb_km),
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_fnb, main = "KDE - Food & Beverage (km)")
plot(sg_owin, add = TRUE)

```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
# IT Services
ppp_it <- acra_62SG_ppp
ppp_it_km <- rescale.ppp(ppp_it, 1000, "km")
kde_it <- density(
  ppp_it_km,
  sigma = bw.diggle(ppp_it_km),
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_it, main = "KDE - IT Services (km)")
plot(sg_owin, add = TRUE)

```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
# Education
ppp_edu <- acra_85SG_ppp
ppp_edu_km <- rescale.ppp(ppp_edu, 1000, "km")
kde_edu <- density(
  ppp_edu_km,
  sigma = bw.diggle(ppp_edu_km),
  edge = TRUE,
  kernel = "gaussian"
)
plot(kde_edu, main = "KDE - Education (km)")
plot(sg_owin, add = TRUE)

```

Notice that output image looks identical to the earlier version, the only changes in the data values (refer to the legend).

**5.3 Working with different automatic badwidth methods**

Beside *bw.diggle()*, there are three other **spatstat** functions can be used to determine the bandwidth, they are: *bw.CvL()*, *bw.scott()*, and *bw.ppl()*.

Let us take a look at the bandwidth return by these automatic bandwidth calculation methods by using the code chunk below.

```{r}
#| eval: false
# Food & Beverage
bw.CvL(ppp_fnb_km)
bw.scott(ppp_fnb_km)
bw.ppl(ppp_fnb_km)
bw.diggle(ppp_fnb_km)

```

```{r}
#| eval: false
# IT Services
bw.CvL(ppp_it_km)
bw.scott(ppp_it_km)
bw.ppl(ppp_it_km)
bw.diggle(ppp_it_km)

```

```{r}
#| eval: false
# Education
bw.CvL(ppp_edu_km)
bw.scott(ppp_edu_km)
bw.ppl(ppp_edu_km)
bw.diggle(ppp_edu_km)

```

**Sample(fnb):**

The code chunk beow will be used to compare the output of using *bw.diggle* and *bw.ppl*methods.

```{r}
#| eval: false
# Food & Beverage KDE (Diggle vs PPL)
kde_fnb_diggle <- density(ppp_fnb_km,
                          sigma = bw.diggle(ppp_fnb_km),
                          edge = TRUE,
                          kernel = "gaussian")

kde_fnb_ppl <- density(ppp_fnb_km,
                       sigma = bw.ppl(ppp_fnb_km),
                       edge = TRUE,
                       kernel = "gaussian")

par(mfrow = c(1, 2))
plot(kde_fnb_diggle, main = "F&B - bw.diggle")
plot(kde_fnb_ppl, main = "F&B - bw.ppl")

```

### **5.4 Working with different kernel methods**

By default, the kernel method used in *density.ppp()* is *gaussian*. But there are three other options, namely: Epanechnikov, Quartic and Dics. The code chunk below will be used to compute three more kernel density estimations by using these three kernel function.

-   **Food & Beverage KDE (different kernels)**

```{r}
#| eval: false
par(mfrow=c(2,2))
plot(density(ppp_fnb_km, sigma=0.4952614, edge=TRUE, kernel="gaussian"),
     main="F&B - Gaussian")
plot(density(ppp_fnb_km, sigma=0.4952614, edge=TRUE, kernel="epanechnikov"),
     main="F&B - Epanechnikov")
plot(density(ppp_fnb_km, sigma=0.4952614, edge=TRUE, kernel="quartic"),
     main="F&B - Quartic")
plot(density(ppp_fnb_km, sigma=0.4952614, edge=TRUE, kernel="disc"),
     main="F&B - Disc")

```

-   **IT Services KDE (different kernels)**

```{r}
#| eval: false
par(mfrow=c(2,2))
plot(density(ppp_it_km, sigma=0.1157808, edge=TRUE, kernel="gaussian"),
     main="IT - Gaussian")
plot(density(ppp_it_km, sigma=0.1157808, edge=TRUE, kernel="epanechnikov"),
     main="IT - Epanechnikov")
plot(density(ppp_it_km, sigma=0.1157808, edge=TRUE, kernel="quartic"),
     main="IT - Quartic")
plot(density(ppp_it_km, sigma=0.1157808, edge=TRUE, kernel="disc"),
     main="IT - Disc")

```

-   **Education KDE (different kernels)**

```{r}
#| eval: false
par(mfrow=c(2,2))
plot(density(ppp_edu_km, sigma=0.1157807, edge=TRUE, kernel="gaussian"),
     main="Edu - Gaussian")
plot(density(ppp_edu_km, sigma=0.1157807, edge=TRUE, kernel="epanechnikov"),
     main="Edu - Epanechnikov")
plot(density(ppp_edu_km, sigma=0.1157807, edge=TRUE, kernel="quartic"),
     main="Edu - Quartic")
plot(density(ppp_edu_km, sigma=0.1157807, edge=TRUE, kernel="disc"),
     main="Edu - Disc")

```

## **6 Fixed and Adaptive KDE**

### **6.1 Computing KDE by using fixed bandwidth**

Compute a KDE layer by your define.

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

# Fixed-bandwidth KDE: 600 m = 0.6 km
kde_fnb_fb <- density.ppp(ppp_fnb_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")
plot(kde_fnb_fb, main = "KDE (Fixed 600 m) – F&B")

```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

kde_it_fb  <- density.ppp(ppp_it_km,  sigma = 0.6, edge = TRUE, kernel = "gaussian")
plot(kde_it_fb, main = "KDE (Fixed 600 m) – it")
```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false
kde_edu_fb <- density.ppp(ppp_edu_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")
plot(kde_edu_fb, main = "KDE (Fixed 600 m) – Education")
```

### **6.2 Computing KDE by using adaptive bandwidth**

Fixed bandwidth method is very sensitive to highly skew distribution of spatial point patterns over geographical units for example urban versus rural. We can solve the problem by using adaptive bandwidth instead.

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

kde_fnb_ab <- adaptive.density(ppp_fnb_km, method="kernel")
plot(kde_fnb_fb, main = "kde_fnb_ab")

```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

kde_it_ab <- adaptive.density(ppp_it_km, method="kernel")
plot(kde_it_ab, main = "kde_it_ab")


```

```{r}
#| fig.height: 6
#| fig.width: 6
#| eval: false

kde_edu_ab <- adaptive.density(ppp_edu_km, method="kernel")
plot(kde_edu_ab, main = "kde_edu_ab")

```

Let's take a look the difference:

```{r}
#| eval: false
par(mfrow=c(1,2))
plot(kde_fnb_fb, main = "Fixed bandwidth")
plot(kde_fnb_ab, main = "Adaptive bandwidth")
```

```{r}
#| eval: false
par(mfrow=c(1,2))
plot(kde_it_fb, main = "Fixed bandwidth")
plot(kde_it_ab, main = "Adaptive bandwidth")
```

```{r}
#| eval: false
par(mfrow=c(1,2))
plot(kde_edu_fb, main = "Fixed bandwidth")
plot(kde_edu_ab, main = "Adaptive bandwidth")
```

## **7 Plotting cartographic quality KDE map**

Convert the im kernal density objects into **SpatRaster** object by using `rast()`of **terra** package.

```{r}

kde_fnb_terra <- readRDS("data/aspatial/kde_rds/kde_fnb_terra.rds")
kde_it_terra  <- readRDS("data/aspatial/kde_rds/kde_it_terra.rds")
kde_edu_terra <- readRDS("data/aspatial/kde_rds/kde_edu_terra.rds")

kde_fnb_terra
kde_it_terra
kde_edu_terra

```

```{r}
#| eval: false
ppp_fnb_km <- rescale.ppp(acra_56SG_ppp, 1000, "km")
ppp_it_km  <- rescale.ppp(acra_62SG_ppp, 1000, "km")
ppp_edu_km <- rescale.ppp(acra_85SG_ppp, 1000, "km")

# Food & Beverage
kde_fnb_fb <- density.ppp(ppp_fnb_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")

# IT Services
kde_it_fb <- density.ppp(ppp_it_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")

# Education
kde_edu_fb <- density.ppp(ppp_edu_km, sigma = 0.6, edge = TRUE, kernel = "gaussian")

# Food & Beverage
kde_fnb_terra <- rast(kde_fnb_fb)
class(kde_fnb_terra)

# IT Services
kde_it_terra <- rast(kde_it_fb)
class(kde_it_terra)

# Education
kde_edu_terra <- rast(kde_edu_fb)
class(kde_edu_terra)

```

### **7.1 Assigning projection systems**

In code chunk below, `crs()` of terra is used to assign the CRS information on *kde_childcareSG_bw_terra* layer.

```{r}
crs(kde_fnb_terra)  <- "EPSG:3414"
crs(kde_it_terra)   <- "EPSG:3414"
crs(kde_edu_terra)  <- "EPSG:3414"
```

```{r}
kde_fnb_terra
kde_it_terra
kde_edu_terra
```

### **7.1.1 Plotting KDE map with tmap**

Display the raster in cartographic quality map using **tmap** package.

```{r}
# Save SpatRaster KDE results
saveRDS(kde_fnb_terra, "data/aspatial/kde_rds/kde_fnb_terra.rds")
saveRDS(kde_it_terra,  "data/aspatial/kde_rds/kde_it_terra.rds")
saveRDS(kde_edu_terra, "data/aspatial/kde_rds/kde_edu_terra.rds")

```

```{r}
#| eval: false

tmap_mode("plot") 
tm_shape(kde_fnb_terra) +
  tm_raster(style = "quantile", n = 7, title = "Density") +
  tm_shape(mp19_subzone_proj) + tm_borders(lwd = 0.4) +
  tm_layout(title = "KDE – Food & Beverage", frame = FALSE)

tm_shape(kde_it_terra) +
  tm_raster(style = "quantile", n = 7, title = "Density") +
  tm_shape(mp19_subzone_proj) + tm_borders(lwd = 0.4) +
  tm_layout(title = "KDE – IT Services", frame = FALSE)

tm_shape(kde_edu_terra) +
  tm_raster(style = "quantile", n = 7, title = "Density") +
  tm_shape(mp19_subzone_proj) + tm_borders(lwd = 0.4) +
  tm_layout(title = "KDE – Education", frame = FALSE)

```

## 

## 8. **First Order SPPA**

We focus on **Downtown Core, Orchard/River Valley, Jurong East, and Tampines** because they are persistent **≥90th-percentile KDE hotspots** across months and across sectors. Together they capture Singapore’s polycentric form—the **CBD**, the premier **retail spine**, and **two regional centres**—so results are comparable while still policy-relevant (street activation, transit access, mixed-use intensity).

## 8.1 **Geospatial data wrangling**

### 8.1.1 Extracting study area

```{r}
#| eval: false

mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)

extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}

mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())

mpsz_sf <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

```

```{r}
write_rds(mpsz_sf, 
          "data/mpsz_cl.rds")
```

```{r}
mpsz_sf <- readr::read_rds("data/mpsz_cl.rds")  
stopifnot(sf::st_crs(mpsz_sf)$epsg == 3414)     

```

The code chunk below will be used to extract the target planning areas.

```{r}
pg <- mpsz_sf %>%
  filter(PLN_AREA_N == "DOWNTOWN CORE")
tm <- mpsz_sf %>%
  filter(PLN_AREA_N == "ORCHARD")
ck <- mpsz_sf %>%
  filter(PLN_AREA_N == "JURONG EAST")
jw <- mpsz_sf %>%
  filter(PLN_AREA_N == "TAMPINES")
```

```{r}
par(mfrow=c(2,2))
plot(st_geometry(pg), main = "DOWNTOWN CORE")
plot(st_geometry(tm), main = "ORCHARD")
plot(st_geometry(ck), main = "JURONG EAST")
plot(st_geometry(jw), main = "TAMPINES")
```

```{r}

pg_owin <- spatstat.geom::as.owin(pg)
tm_owin <- spatstat.geom::as.owin(tm)
ck_owin <- spatstat.geom::as.owin(ck)
jw_owin <- spatstat.geom::as.owin(jw)


```

```{r}
acra_56_ppp <- spatstat.geom::as.ppp(acra_56)
acra_62_ppp <- spatstat.geom::as.ppp(acra_62)
acra_85_ppp <- spatstat.geom::as.ppp(acra_85)

# F&B
fnb_pg_ppp <- acra_56_ppp[pg_owin]; fnb_tm_ppp <- acra_56_ppp[tm_owin]
fnb_ck_ppp <- acra_56_ppp[ck_owin]; fnb_jw_ppp <- acra_56_ppp[jw_owin]

# IT
it_pg_ppp  <- acra_62_ppp[pg_owin]; it_tm_ppp  <- acra_62_ppp[tm_owin]
it_ck_ppp  <- acra_62_ppp[ck_owin]; it_jw_ppp  <- acra_62_ppp[jw_owin]

# Education
edu_pg_ppp <- acra_85_ppp[pg_owin]; edu_tm_ppp <- acra_85_ppp[tm_owin]
edu_ck_ppp <- acra_85_ppp[ck_owin]; edu_jw_ppp <- acra_85_ppp[jw_owin]

```

```{r}
fnb_pg_km <- spatstat.geom::rescale.ppp(fnb_pg_ppp, 1000, "km")
fnb_tm_km <- spatstat.geom::rescale.ppp(fnb_tm_ppp, 1000, "km")
fnb_ck_km <- spatstat.geom::rescale.ppp(fnb_ck_ppp, 1000, "km")
fnb_jw_km <- spatstat.geom::rescale.ppp(fnb_jw_ppp, 1000, "km")

it_pg_km  <- spatstat.geom::rescale.ppp(it_pg_ppp,  1000, "km")
it_tm_km  <- spatstat.geom::rescale.ppp(it_tm_ppp,  1000, "km")
it_ck_km  <- spatstat.geom::rescale.ppp(it_ck_ppp,  1000, "km")
it_jw_km  <- spatstat.geom::rescale.ppp(it_jw_ppp,  1000, "km")

edu_pg_km <- spatstat.geom::rescale.ppp(edu_pg_ppp, 1000, "km")
edu_tm_km <- spatstat.geom::rescale.ppp(edu_tm_ppp, 1000, "km")
edu_ck_km <- spatstat.geom::rescale.ppp(edu_ck_ppp, 1000, "km")
edu_jw_km <- spatstat.geom::rescale.ppp(edu_jw_ppp, 1000, "km")

```

```{r}
#| fig.height: 6
#| fig.width: 6


par(mfrow=c(2,2))
plot(unmark(fnb_pg_km), main="F&B — Downtown Core")
plot(unmark(fnb_tm_km), main="F&B — Orchard")
plot(unmark(fnb_ck_km), main="F&B — Jurong East")
plot(unmark(fnb_jw_km), main="F&B — Tampines")
par(mfrow=c(1,1))

```

```{r}
#| fig.height: 6
#| fig.width: 6


par(mfrow=c(2,2))
plot(unmark(it_pg_km), main="it — Downtown Core")
plot(unmark(it_tm_km), main="it — Orchard")
plot(unmark(it_ck_km), main="it — Jurong East")
plot(unmark(it_jw_km), main="it — Tampines")
par(mfrow=c(1,1))

```

```{r}
#| fig.height: 6
#| fig.width: 6

par(mfrow=c(2,2))
plot(unmark(edu_pg_km), main="edu — Downtown Core")
plot(unmark(edu_tm_km), main="edu — Orchard")
plot(unmark(edu_ck_km), main="edu — Jurong East")
plot(unmark(edu_jw_km), main="edu — Tampines")
par(mfrow=c(1,1))

```

### 8.1.2 **Clark and Evans Test**

```{r}
run_CE <- function(ppp_obj, alt = "two.sided", nsim = 999){
  ce <- clarkevans.test(ppp_obj, correction="none",
                        clipregion=NULL, alternative=alt, nsim=nsim)
  print(ce); invisible(ce)}
```

###  F&B Industry

```{r}
run_CE(fnb_pg_km)  # Downtown Core
run_CE(fnb_tm_km)  # Orchard
run_CE(fnb_ck_km)  # Jurong East
run_CE(fnb_jw_km)  # Tampines
```

###  IT Industry

```{r}
# --- IT ---
run_CE(it_pg_km)# Downtown Core
run_CE(it_tm_km)# Orchard
run_CE(it_ck_km)# Jurong East
run_CE(it_jw_km) # Tampines

```

### Education Industry

```{r}
# --- Education ---
run_CE(edu_pg_km)# Downtown Core
run_CE(edu_tm_km)# Orchard
run_CE(edu_ck_km)# Jurong East
run_CE(edu_jw_km) # Tampines
```

### 8.1.3 **Computing KDE surfaces by planning area**

The code chunk below will be used to compute the KDE of these four planning area. ***bw.diggle*** method is used to derive the bandwidth of each

```{r}
#| fig.width: 11
#| fig.height: 9
#| dpi: 200

# --- F&B ---
par(mfrow=c(2,2), mar=c(3.5,3.5,3,5), cex=1.15)
plot(density(edu_pg_km,  sigma=bw.diggle(edu_pg_km),  edge=TRUE),
     main="Edu — Downtown Core",  ribargs=list(cex.axis=1.2))
plot(density(edu_tm_km,  sigma=bw.diggle(edu_tm_km),  edge=TRUE),
     main="Edu — Orchard",        ribargs=list(cex.axis=1.2))
plot(density(edu_ck_km,  sigma=bw.diggle(edu_ck_km),  edge=TRUE),
     main="Edu — Jurong East",    ribargs=list(cex.axis=1.2))
plot(density(edu_jw_km,  sigma=bw.diggle(edu_jw_km),  edge=TRUE),
     main="Edu — Tampines",       ribargs=list(cex.axis=1.2))
par(mfrow=c(1,1))


# --- IT ---
par(mfrow=c(2,2))
plot(density(it_pg_km,  sigma=bw.diggle(it_pg_km),  edge=TRUE, kernel="gaussian"),
     main="IT — Downtown Core")
plot(density(it_tm_km,  sigma=bw.diggle(it_tm_km),  edge=TRUE, kernel="gaussian"),
     main="IT — Orchard")
plot(density(it_ck_km,  sigma=bw.diggle(it_ck_km),  edge=TRUE, kernel="gaussian"),
     main="IT — Jurong East")
plot(density(it_jw_km,  sigma=bw.diggle(it_jw_km),  edge=TRUE, kernel="gaussian"),
     main="IT — Tampines")
par(mfrow=c(1,1))

# --- Education ---
par(mfrow=c(2,2))
plot(density(edu_pg_km, sigma=bw.diggle(edu_pg_km), edge=TRUE, kernel="gaussian"),
     main="Edu — Downtown Core")
plot(density(edu_tm_km, sigma=bw.diggle(edu_tm_km), edge=TRUE, kernel="gaussian"),
     main="Edu — Orchard")
plot(density(edu_ck_km, sigma=bw.diggle(edu_ck_km), edge=TRUE, kernel="gaussian"),
     main="Edu — Jurong East")
plot(density(edu_jw_km, sigma=bw.diggle(edu_jw_km), edge=TRUE, kernel="gaussian"),
     main="Edu — Tampines")
par(mfrow=c(1,1))

```

## 9.  **2 nd Order Spatial Point Patterns Analysis Methods**

```{r}
set.seed(1234)
```

## 9.1 **Analysing Spatial Point Process Using G-Function**

```{r}
pa_dc <- mpsz_sf %>% dplyr::filter(PLN_AREA_N == "DOWNTOWN CORE")
pa_or <- mpsz_sf %>% dplyr::filter(PLN_AREA_N == "ORCHARD")
pa_je <- mpsz_sf %>% dplyr::filter(PLN_AREA_N == "JURONG EAST")
pa_tm <- mpsz_sf %>% dplyr::filter(PLN_AREA_N == "TAMPINES")

w_dc <- spatstat.geom::as.owin(pa_dc)
w_or <- spatstat.geom::as.owin(pa_or)
w_je <- spatstat.geom::as.owin(pa_je)
w_tm <- spatstat.geom::as.owin(pa_tm)

```

Arccording the task of caculative the function is big,so we select one to clarify:

```{r}
ex_file <- file.path(out_dir, "FNB_56__Downtown_Core.rds")
  plot(obj$envelopes$G, main = "G envelope")
```

```{r}
 plot(obj$envelopes$F, main = "F envelope")
```

```{r}
plot(obj$envelopes$K, . - r ~ r, xlab="d (m)", ylab="K(d)-r", main="K envelope"); abline(h=0, lty=2)
```

```{r}
plot(obj$envelopes$L, . - r ~ r, xlab="d (m)", ylab="L(r)-r", main="L envelope"); abline(h=0, lty=2)
```

```{r}
stkde_fnb <- readRDS("data/aspatial/kde_rds/STKDE_FNB_2025H1.rds")
par(mfcol=c(2,3)); for(i in 1:6) plot(stkde_fnb, i, fix.range=TRUE,
                                      main=paste("F&B STKDE — 2025-", sprintf("%02d",i), sep=""))
par(mfcol=c(1,1))
```

```{r}
#| fig.heigh: 12
#| fig.width: 12
#| warning: false
fnb_2025 <- acra_sf_proj %>%
  filter(ssic_2d == "56",
         reg_date >= as.Date("2025-01-01"),
         reg_date <= as.Date("2025-06-30")) %>%
  mutate(month_num = month(reg_date))
fnb_ppp <- spatstat.geom::as.ppp(fnb_2025 %>% select(month_num))
get_win <- function(pa) {
  mpsz_sf %>% filter(PLN_AREA_N == pa) %>%
    st_union() %>% spatstat.geom::as.owin()
}

areas <- c("DOWNTOWN CORE","ORCHARD","JURONG EAST","TAMPINES")
for (pa in areas) {
  win <- get_win(pa)
  fnb_pa <- fnb_ppp[win]                                  
  stkde <- sparr::spattemp.density(fnb_pa)                
  out <- paste0("data/aspatial/kde_rds/STKDE_FNB_",
                gsub(" ","",pa), "_2025H1.rds")
  saveRDS(stkde, out)
  par(mfcol = c(2,3), mar = c(2,2,2,2))
  for (i in 1:6) {
    plot(stkde, i, override.par=FALSE, fix.range=TRUE,
         main = paste0("F&B STKDE — ", pa, " (", sprintf("2025-%02d", i), ")"))
    plot(win, add = TRUE)
  }
  par(mfcol = c(1,1))
}

```
