---
title: "Hands-on Exercise 10: Local Measures of Spatial Autocorrelation"
author: "BingBing"
date: "September 18, 2025"
date-modified: "last-modified"
---

## 10.1 Overview — Local Measures of Spatial Autocorrelation (LMSA)

LMSA assess how each observation relates to its surrounding neighbors, yielding location-specific scores rather than a single map-wide statistic. They share intuition with global measures and, in some cases, the global metric can be decomposed into local components—e.g., LISA (Local Indicators of Spatial Association). In addition to LISA, Getis–Ord 𝐺 𝑖 G i ​

is introduced as an alternative LMSA that offers complementary hot/cold-spot insights.

By the end of this exercise you will be able to:

1.  import geospatial data with sf;

2.  import CSV data with readr;

3.  perform relational joins with dplyr;

4.  compute LISA (to detect local clusters and outliers) with sfdep;

5.  compute Getis–Ord 𝐺 𝑖 (to detect hot/cold spots) with sfdep;

6.  visualize the outputs with tmap.

## 10.2 Getting Started 10.2.1 The analytical question

In spatial policy, a central goal is equitable spatial distribution of development. We ask: (i) Are development levels evenly distributed? If no, (ii) is there spatial clustering? If yes, (iii) where are the clusters? The case study examines GDP per capita in Hunan Province, PRC.

### 10.2.2 The Study Area and Data

-   Area: Hunan Province at county level.

-   Datasets: (1) county administrative boundary shapefile; (2) Hunan_2012.csv containing selected 2012 development indicators.

### 10.2.3 Setting the Analytical Tools

Ensure the following R packages are installed and loaded:

-   sf (import/handle geospatial data),

-   tidyverse (wrangle attribute data),

-   sfdep (compute spatial weights; global and local spatial autocorrelation),

-   tmap (produce cartographic-quality choropleth maps).

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## **10.3 Getting the Data Into R Environment**

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

### **10.3.1 Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

In the code chunk above, `st_transform()` is used to transform Hunan geospatial data from WGS 84 to UTM zone 50N (i.e. EPSG: 32650).

### **10.3.2 Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R data frame class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### **10.3.3 Performing relational join**

The code chunk below will be used to update the attribute table of *hunan’s* sf data frame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### **10.3.4 Visualising Regional Development Indicator**

In the code chunks below, tmap functions are used:

-   to build two choropleth maps by using equal interval (i.e equal) and quantile (i.e. quantile) classification methods, and

-   to plot both maps next to each other by using `tmap_arrange()`.

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## **10.4 Local Indicators of Spatial Association (LISA)**

LISA provides **location-specific** statistics to detect **clusters** (high–high, low–low) and **outliers** (high–low, low–high) in a spatial variable. Here, we use **local Moran’s I** to identify where GDP per capita in Hunan (2012) departs from what pure spatial randomness would produce, highlighting counties whose values are unusually similar to—or different from—their neighbors.

### **10.4.1 Computing Contiguity Spatial Weights**

First build neighborhood structure, then convert it to weights. Use **`sfdep::st_contiguity()`** to create a first-order contiguity neighbors list; by default `queen = TRUE` (shared edge **or** corner qualifies). Next, pass the neighbors list to **`sfdep::st_weights()`** to obtain polygon weights.

Next, [`st_weights()`](https://sfdep.josiahparry.com/reference/st_weights) is used to calculate polygon spatial weights from the *nb* list.

`st_weights()` provides three arguments, they are:

-   nb: A neighbor list object as created by `st_neighbors()`.

-   style: Default “W” for row standardized weights. This value can also be “B”, “C”, “U”, “minmax”, and “S”. B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   allow_zero: If TRUE, assigns zero as lagged value to zone without neighbors.

```{r}
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1) 
```

To reveal the neighbour list, code chunk below is used.

```{r}
summary(wm_q$nb)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbour.

### 10.4.2 Row-standardised weights

Row-standardising (**style = "W"**) gives each neighbor equal weight $1/k$ (where $k$ is the number of neighbors), so a unit’s spatial lag is the **average** of its neighbors’ values. This is simple and comparable across units, but **edge units** (with fewer neighbors) may have lags based on scant information, which can distort perceived autocorrelation. Simplicity wins here, so we keep `"W"`, noting that alternatives like **`"B"`** (binary, unstandardised) exist.

### What the code chunk teaches

-   **Required input:** `nb2listw()` takes an **`nb`** (neighbors list) object.

-   **Two key args:**

    -   **`style`** – weighting scheme:

        -   `"W"` row-standardised (rows sum to 1);
        -   `"B"` binary (0/1 links, unstandardised);
        -   `"C"` globally standardised (all weights sum to *n*);
        -   `"U"` = `"C"` divided by each unit’s neighbor count (rows sum to 1);
        -   `"minmax"` rescaled to $[0,1]$;
        -   `"S"` Tiefelsdorf variance-stabilising (sums to *n*).

    -   **`zero.policy`** – handling **isolates** (no neighbors). If `TRUE`, inserts zero-length weight vectors so the **spatial lag = 0** for those units—sometimes sensible, sometimes not; use with care.

### **10.4.3 Computing local Moran’s I**

To compute local Moran’s I, the [`local_moran()`](https://sfdep.josiahparry.com/reference/local_moran) function of **sfdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran’s I of *GDPPC2012* at the county level.

```{r}
set.seed(1234)
```

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

`local_moran()` function returns a matrix of values whose columns are:

-   Ii: the local Moran’s I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

```{r}
glimpse(lisa)
```

#### 10.4.3.1 Mapping the local Moran’s I

#### 10.4.3.2 Mapping local Moran’s I values

Using choropleth mapping functions of **tmap** package, we can plot the local Moran’s I values by using the code chinks below.

```{r}
#| warning: false
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```

#### 10.4.3.3 Mapping local Moran’s I p-values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran’s I p-values by using functions of **tmap** package.

```{r}
#| warning: false
tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```

#### 10.4.3.4 Mapping both local Moran’s I values and p-values

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
#| warning: false
ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Moran's I",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Moran's I of GDPPC (Queen's method)")

p_ii.map <- tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")

tmap_arrange(ii.map, p_ii.map, asp=1, ncol=2)
```

### **10.4.4 Preparing & Visualising LISA**

This section explains how to turn local autocorrelation results into intuitive maps. A **LISA Cluster Map** highlights statistically significant locations by type: **High–High (hot spots)**, **Low–Low (cold spots)**, and **outliers** (**High–Low**, **Low–High**). The usual workflow begins with a **Moran scatterplot**, which plots each unit’s **standardized value** (x-axis) against its **spatial lag** (y-axis). The plot’s slope equals local Moran’s *I* and its four quadrants correspond to HH, LL, HL, and LH patterns; significance comes from local tests.

#### 10.4.4.1 Plotting Moran scatterplot

A Moran Scatterplot is a graphical tool in spatial analysis that visualizes the relationship between a variable’s values and the average values of its neighboring locations (its spatial lag) to identify spatial autocorrelation. By plotting standardized values of the variable on the x-axis and its spatial lag on the y-axis, the plot’s slope represents Moran’s I, revealing patterns of high-high or low-low value clusters and high-low or low-high value outliers.

In the code chunk below, [`st_lag()`](https://sfdep.josiahparry.com/reference/st_lag) of **sfdep** package is used to calculates the spatial lag of GDPPC given a neighbor (i.e. nb) and weights list (i.e. wt).

```{r}
lisa <- lisa %>%
  mutate(lag_GDPPC = st_lag(
    GDPPC, nb, wt),
    .before = 1) %>%
  unnest(lag_GDPPC)
```

Then, the code chunk below is used to plot the Moran Scatterplot of GDPPC against its spatial lag version as shown in the code cunk below.

```{r}
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC)) +
  geom_point() +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "red") +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot") +
  theme_minimal()
```

The Moran Scatterplot above is not very intuitive. To support effective visual communication, the code chunk below is used.

```{r}
#| warning: false
ggplot(data = lisa, 
       aes(x = GDPPC, 
           y = lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "GDPPC",
       y = "Spatial Lag of GDPPC",
       title = "Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

#### 10.4.4.2 Plotting Moran scatterplot with standardised variable

First we will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
lisa <- lisa %>%
  mutate(z_GDPPC = scale(GDPPC),
         z_lag_GDPPC = scale(lag_GDPPC),
         .before = 1)
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
#| warning: false
ggplot(data = lisa, 
       aes(x = z_GDPPC, 
           y = z_lag_GDPPC, 
           color = mean)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", 
              se = FALSE, 
              color = "black") +
  geom_hline(yintercept=mean(lisa$z_lag_GDPPC), lty=2) + 
    geom_vline(xintercept=mean(lisa$z_GDPPC), lty=2) +
  scale_color_manual(
    values = c("High-High" = "red", 
               "Low-Low" = "blue",
               "Low-High" = "lightblue", 
               "High-Low" = "pink")) +
  labs(x = "Standardised GDPPC",
       y = "Standardised Spatial Lag of GDPPC",
       title = "Standardised Moran Scatterplot with LISA Quadrants") +
  theme_minimal()
```

#### 10.4.4.3 Preparing LISA map classes

The code chunks below show the steps to prepare a LISA cluster map.

Next, derives the spatially lagged variable of interest (i.e. GDPPC) and centers the spatially lagged variable around its mean.

This is follow by centering the local Moran’s around the mean.

Next, we will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05       
```

#### 10.4.4.4 Plotting and visualising LISA map

By default, the LISA classes are stored in *mean*, *median* and *pysal* fields of lisa sf data frame. To prepare the LISA map, we will derive a new field called LISA-cluster from either one of these three field by using the code chunk below.

```{r}
lisa <- lisa %>%
  mutate(
    LISA_cluster = ifelse(
      p_ii < 0.05,
      as.character(mean), 
      "Insignificant"),
    LISA_cluster = factor(
      LISA_cluster,
      levels = c("Insignificant", "Low-Low", "Low-High", "High-Low", "High-High")
    )
  )
```

Now, we can plot the LISA map by using the code chunk below.

```{r}
lisa_map <- tm_shape(lisa) + 
  tm_polygons(
    fill = "LISA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "blue",        # Low-Low
        "lightblue",   # Low-High
        "pink",        # High-Low
        "red"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "LISA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("Local Moran's I Clusters (p < 0.05)")
lisa_map
```

For effective interpretation, it is better to plot both the local Moran’s I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

We can also include the local Moran’s I map and p-value map as shown below for easy comparison.

```{r}
#| warning: false

tmap_mode("plot") 
tmap_arrange(ii.map, lisa_map, ncol = 2, asp = 1)
```

## **10.5 Hot Spots and Cold Spots Analysis (HCSA)**

Hot spots and cold spots analysis is a technique used in spatial statistics to identify statistically significant clusters of high values (hot spots) and low values (cold spots) of a geographically referenced attribute (i.e. GDPPC). These analyses help visualize and understand spatial patterns, revealing areas where data points are clustered more densely than would be expected by random chance.

Figure below shows the hot spots and cold spots of GDP per capita of Hunan Provice, People Republic of China.

The term hot spots refer to areas where high values are concentrated and surrounded by other high values. It has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015). For example, a hot spot analysis might reveal clusters of high GDP per capita in a province, indicating areas with significantly higher GDP per capita than other parts of the province.

On the other hands, the term cold spots refer to areas where low values are concentrated and surrounded by other low values. An example could be clusters of low GDP per capita, indicating areas where GDP per capita is consistently low.

HCSA uses statistical methods, like the Getis-Ord Gi\* statistic, to determine if the observed clustering is statistically significant or could have occurred randomly.

The analysis consists of four steps:

-   Deriving spatial weight matrix

-   Computing Gi\* statistics

-   Mapping Gi\* statistics

-   Communicating the analysis results

### **10.5.1 Deriving distance weight matrix**

Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord Gi\* statistics, we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

-   fixed distance weight matrix,

-   adaptive distance weight matrix, and

-   inverse distance weights.

#### 10.5.1.1 Computing fixed distance weights

In this section, you will learn how to compute fixed distance weight matrix by using `st_dist_band()` and `st_weights()` of sfdep package.

Firstly, we need to determine the upper limit for distance band by using the steps below:

```{r}
#| warning: false
ct <- critical_threshold(st_geometry(hunan))
ct
```

The summary report shows that the largest first nearest neighbour distance is 60.80 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

Using the critical threshold value computed above, code chunk below is used to compute the fixed distance weight.

```{r}
#| warning: false
hunan_fdw <- hunan %>% 
  mutate(nb = include_self(
    st_dist_band(
      st_geometry(geometry), 
                 upper = ct)),
    wt = st_weights(nb, 
                      style = "W"),
    .before = 1) 
```

#### 10.5.1.2 Computing adaptive distance weights

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smooth the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

In this section, you will learn how to compute fixed distance weight matrix by using `st_dist_band()` and `st_weights()` of sfdep package.

```{r}
#| warning: false
hunan_adw <- hunan %>%
  mutate(nb = include_self(
    st_knn(
      st_geometry(geometry),
      k = 6)),
    wt = st_weights(
      nb, style = "W"),
    .before = 1)
```

#### 10.5.1.3 Computing adaptive distance weights

### **10.5.2 Computing local Gi\***

Local Gi\* (pronounced “G-i-star”) is a Getis-Ord statistic used in spatial analysis to identify statistically significant hot spots and cold spots of high or low attribute values within a dataset. It measures the spatial clustering of values by comparing the total value within a specified neighborhood to the global total, producing a Z-score and p-value for each location to determine the likelihood that the observed pattern is not due to random chance.

In this sub-section, you will gain hands-on experience on performing Gi\* analysis by using `local_gstar_perm()` of sfdep package.

```{r}
HCSA_fdw <- hunan_fdw %>%
  mutate(
    gistar = local_gstar_perm(
      GDPPC, nb, wt, nsim = 99),
    .before = 1) %>%
  unnest(gistar)
```

### **10.5.3 Preparing and Visualising HCSA Map**

#### 10.5.3.1 Mapping Gi\* with fix distance weights

In the code chunk below, thematic mapping functions of tmaps are used to prepare a choropleth map showing the distribution of Gi\* values.

```{r}
#| warning: false
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 6,
                values = "brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "Gi*",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Gi* of GDPPC (Fixed Bandwidth d = 60799.91m)")
```

#### 10.5.3.2 Mapping local Gi\* p-values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunk below plots a choropleth map of p-values of Gi\* by using functions of **tmap** package.

```{r}
#| warning: false
tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "simulated p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (Fixed distance)")
```

#### 10.5.3.3 Mapping both Gi\* values and p-values

For effective interpretation, it is better to plot both the Gi\* values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
#| warning: false
Gi_star_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "gi_star",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "-brewer.rd_bu"),
              fill.legend = tm_legend(
                title = "local Gi*",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Local Gi* of GDPPC")

p_values_map <- tm_shape(HCSA_fdw) +
  tm_polygons(fill = "p_sim", 
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.001, 0.01, 0.05, 0.1, 1),
                values = "-brewer.reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_in("left", "bottom")
    )) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of local Gi* of GDPPC (fixed distance)")

tmap_arrange(Gi_star_map, p_values_map, asp=1, ncol=2)
```

#### 10.5.3.4 Plotting and visualising HCSA Map

By default, the HCSA classes are stored in *cluster* field of *HCSA_fdw* sf data frame. To prepare the HSCA map, we will derive a new field called *HCSA_cluster* from *cluster* field by using the code chunk below.

```{r}
HCSA_fdw <- HCSA_fdw %>%
  mutate(HCSA_cluster = case_when(
    p_sim > 0.05 ~ "Insignificant",
    p_sim <= 0.05 & cluster == "High" ~ "Hot spot",
    p_sim <= 0.05 & cluster == "Low"  ~ "Cold spot",
    TRUE ~ "Other"),
    HCSA_cluster = factor(
      HCSA_cluster,
      levels = c("Insignificant", "Hot spot", "Cold spot")
    ),
    .before = 1
  )

```

Now, we can plot the HCSA map by using the code chunk below.

```{r}
HCSA_map <- tm_shape(HCSA_fdw) + 
  tm_polygons(
    fill = "HCSA_cluster",
    fill.scale = tm_scale_categorical(
      values = c(
        "grey80",      # Insignificant
        "red",        # Low-Low
        "blue"          # High-High
      )
    ),
    fill.legend = tm_legend(
      title = "HSCA Cluster",
      position = tm_pos_in("left", "bottom"))
  ) +
  tm_borders() +
  tm_title("HCSA Clusters (p < 0.05)")
HCSA_map
```

For effective interpretation, it is better to plot both the local Gi\* values map and its corresponding HCSA map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
#| warning: false
tmap_arrange(Gi_star_map, HCSA_map, asp=1, ncol=2)
```

**💡 Observation:**

The Gi\* (local) map and the HCSA cluster map (p \< 0.05) both show a contiguous hot‐spot belt in the east–central counties, indicating a statistically significant concentration of high GDP per capita there.

Several cold spots (significantly low GDP per capita) appear in the north-west and south-west, with a small one near the center; most remaining counties are not significant.

Overall pattern: local clustering and spatial heterogeneity—a clear east (high) vs. west (low) contrast rather than a uniform surface, consistent with positive local spatial autocorrelation.
