---
title: "Spatio-Temporal Point Patterns Analysis"
---

## **6.1 Overview**

A spatio-temporal point process records events by **where** they happen and **when** they occur—for example, disease cases, species sightings, or fire incidents. As more datasets come with both coordinates and timestamps, it’s increasingly useful to explore their patterns in space and time.\

In this chapter, we walk through a practical workflow in R to analyze such patterns using a real-world example: forest-fire events in **Kepulauan Bangka Belitung, Indonesia** from **2023-01-01** to **2023-12-31**.

## **6.2 Learning Outcome**

By the end, you should be able to:

-   Judge whether the observed fire events behave like random points in **space** and in **space–time**.

-   If they are **not** random, pinpoint **where** hotspots appear and **when** clustering intensifies, and present clear evidence with maps and simple statistics.

### **6.2.1 The research questions**

1.  Are forest-fire locations in Kepulauan Bangka Belitung **spatially** and **spatio-temporally** independent?\
2.  If **not**, **where** and **when** do the fires tend to **cluster**?

## **6.3 The data**

-   *forestfires* (CSV): point records of forest fires detected by the MODIS sensor (from FIRMS).\
    For this exercise we only keep events within **Kepulauan Bangka Belitung**.

-   *Kepulauan_Bangka_Belitung* (ESRI Shapefile): sub-district (**kelurahan**) boundaries for Kepulauan Bangka Belitung (extracted from Indonesia’s national boundary dataset).

## **6.4 Installing and Loading the R packages**

For the purpose of this study, six R packages will be used. They are:

| **Package** | Data type | Purpose in this study | Core functions (examples) |
|------------------|------------------|------------------|------------------|
| [**sf**](https://r-spatial.github.io/sf/) | Vector (points/lines/polygons) | Load kelurahan boundaries, convert fire CSV to points, reproject, spatial join | `st_read()`, `st_as_sf()`, `st_transform()`, `st_intersection()` |
| [**raster**](https://rspatial.org/raster/) | Raster (grids) | Store/compute density or risk surfaces and sample values | `raster()`, `crop()`, `mask()`, `extract()` |
| [**spatstat**](https://cran.r-project.org/web/packages/spatstat/index.html) | Point pattern stats | Test CSR and clustering; analyze cross-type interactions | `ppp()`, `Kest()`, `Lest()`, `Kcross()`, `Lcross()` |
| [**sparr**](https://tilmandavies.github.io/sparr/index.html) | Kernel density & risk | Build intensity and relative-risk surfaces; basic space-time | `bivar()`, `risk()`, `spattemp.dens()` |
| [**tmap**](https://r-tmap.github.io/tmap/) | Mapping | Make static/interactive choropleths and small multiples | `tm_shape()`, `tm_polygons()`, `tm_borders()`, `tm_facets()`, `tmap_mode()` |
| [**tidyverse**](https://www.tidyverse.org/)  | Wrangling | Import, clean, reshape, and summarize tabular data | `read_csv()`, `mutate()`, `filter()`, `pivot_longer()/wider()` |

### 

```{r}
pacman::p_load(sf, terra, spatstat, sparr, tmap, tidyverse)
```

## **6.5 Importing and Preparing Study Area**

### **6.5.1 Importing study area**

Code chunk below is used import study area (i.e. Kepulauan Bangka Belitung) into R environment.

```{r}
kbb <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") 
```

The revised code chunk.

```{r}
kbb_sf <- st_read(dsn="data/rawdata",
               layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

```{r}
#| echo: false

kbb_sf <- st_as_sf(kbb_sf)                              
kbb_sf <- st_cast(kbb_sf, "POLYGON", warn = FALSE)
kbb_sf <- kbb_sf[which.max(st_area(kbb_sf)), , drop = FALSE]
```



### **6.5.2 Converting OWIN**

Next, `as.owin()` is used to convert *kbb* into an **owin** object.

```{r}
kbb_owin <- spatstat.geom::as.owin(kbb_sf)
kbb_owin

```

window: polygonal boundary
enclosing rectangle: [512066.8, 705559.4] x [9655398, 9834006] units
```

Next, `class()` is used to confirm if the output is indeed an owin object.

```{r}
class(kbb_owin)
```

## **6.6 Importing and Preparing Forest Fire data**

Next, we will import the forest fire data set (i.e. *forestfires.csv*) into R environment.

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
                       crs = 4326) %>%
  st_transform(crs = 32748)
```
```{r}
#| echo: false
if (st_crs(fire_sf) != st_crs(kbb_sf)) {
  fire_sf <- st_transform(fire_sf, st_crs(kbb_sf))
}
fire_sf <- st_intersection(fire_sf, kbb_sf)
```

Because **ppp** object only accept numerical or character as mark. The code chunk below is used to convert data type of *acq_date* to numeric.

```{r}
fire_sf <- fire_sf %>% 
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date, 
                           label = TRUE, 
                           abbr = FALSE))
```

## **6.7 Visualising the Fire Points**

### **6.7.1 Overall plot**

```{r}
kbb_owin <- spatstat.geom::as.owin(kbb_sf)

tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots()
```

### **6.7.2 Visuaising geographic distribution of forest fires by month**

```{r}
tm_shape(kbb_sf)+
  tm_polygons() +
tm_shape(fire_sf) +
  tm_dots(size = 0.) +
tm_facets(by="Month_fac", 
            free.coords=FALSE, 
            drop.units = TRUE)
```

## **6.8 Computing STKDE by Month**

In this section, you will learn how to compute STKDE by using [`spattemp.density()`](https://tilmandavies.github.io/sparr/reference/spattemp.density.html) of **sparr** package. Before using the function, it is highly recommended you read the function’s reference guide in detail in order to understand the input data requirements and the output object generated.

### **6.8.1 Extracting forest fires by month**

The code chunk below is used to remove the unwanted fields from *fire_sf* sf data.frame. This is because `as.ppp()` only need the mark field and geometry field from the input sf data.frame.

```{r}
fire_month <- fire_sf %>% 
  select(Month_num)
```

### **6.8.2 Creating ppp**

The code chunk below is used to derive a **ppp** object called *fire_month* from *fire_month*sf data.frame.

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

The code chunk below is used to check the output is in the correct object class.

```{r}
summary(fire_month_ppp)
```

Next, we will check if there are duplicated point events by using the code chunk below.

```{r}
any(duplicated(fire_month_ppp))
```

### **6.8.3 Including Owin object**

The code chunk below is used to combine *origin_am_ppp* and *am_owin* objects into one.

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

As a good practice, `plot()` is used to plot *ff_owin* so that we can examine the correctness of the output object.

```{r}
#| echo: false
#| fig.height: 9
#| fig.width: 9

plot(fire_month_owin)
```

### **6.8.4 Computing Spatio-temporal KDE**

Next, `spattemp.density()` of sparr package is used to compute the STKDE.

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### **6.8.5 Plotting the spatio-temporal KDE object**

In the code chunk below, `plot()` of R base is used to the KDE for between July 2023 - December 2023.

```{r}
#| echo: false
#| fig.height: 9
#| fig.width: 9

tims <- c(7,8,9,10,11,12)
par(mfcol=c(2,3))
for(i in tims){ 
  plot(st_kde, i, 
       override.par=FALSE, 
       fix.range=TRUE, 
       main=paste("KDE at month",i))
}
```

## **6.9 Computing STKDE by Day of Year**

In this section, you will learn how to computer the STKDE of forest fires by day of year.

### **6.9.1 Creating ppp object**

In the code chunk below, DayofYear field is included in the output ppp object.

```{r}
fire_yday_ppp <- fire_sf %>% 
  select(DayofYear) %>%
  as.ppp()
```

### **6.9.2 Including Owin object**

Next, code chunk below is used to combine the ppp object and the owin object.

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### **6.9.3**

```{r}
kde_yday <- spattemp.density(
  fire_yday_owin)
summary(kde_yday)
```

```{r}
plot(kde_yday)
```

## **6.10 Computing STKDE by Day of Year: Improved method**

One of the nice function provides in sparr package is BOOT.spattemp(). It support bandwidth selection for standalone spatiotemporal density/intensity based on bootstrap estimation of the MISE, providing an isotropic scalar spatial bandwidth and a scalar temporal bandwidth.

Code chunk below uses BOOT.spattemp() to determine both the spatial bandwidth and the scalar temporal bandwidth.

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin) 
```

### **6.10.1 Computing spatio-temporal KDE**

Now, the STKDE will be derived by using h and lambda values derive in previous step.

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 9000,
                             lambda = 19)
summary(kde_yday)
```

### **6.10.2 Plotting the output spatio-temporal KDE**

Last, plot() of sparr package is used to plot the output as shown below.

```{r}
plot(kde_yday)
```
