---
title: "Hands on Exercise 12 - Geographical Segmentation with Spatially Constrained Clustering Techniques"
---

## 12.1 Overview

In this exercise, I learned how to delineate **homogeneous regions** using geographically referenced multivariate data. Two types of analysis were carried out:\
1. **Hierarchical cluster analysis**\
2. **Spatially constrained cluster analysis**

### 12.1.1 Learning Outcome

By completing this hands-on exercise, I was able to:\
- Convert GIS polygon data into R’s **sf** data frame\
- Transform an **sf** object into a **SpatialPolygonDataFrame**\
- Perform hierarchical clustering using **hclust()**\
- Apply spatially constrained clustering with **skater()**\
- Visualize the outputs with **ggplot2** and **tmap**

## 12.2 Getting Started

In geobusiness and spatial policy analysis, it is common to delineate markets or planning areas into homogeneous regions using multivariate data.\
In this exercise, the focus is on **Shan State, Myanmar**, where multiple ICT indicators (Radio, Television, Landline phone, Mobile phone, Computer, and Internet at home) are used to identify homogeneous regions.

## **12.3 The data**

Two data sets will be used in this study. They are:

-   Myanmar Township Boundary Data (i.e. *myanmar_township_boundaries*) : This is a GIS data in ESRI shapefile format. It consists of township boundary information of Myanmar. The spatial data are captured in polygon features.

-   *Shan-ICT.csv*: This is an extract of [**The 2014 Myanmar Population and Housing Census Myanmar**](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level.

Both data sets are download from [Myanmar Information Management Unit (MIMU)](http://themimu.info/)

### 12.3.1 Installing and Loading R Packages

Before starting the analysis, it is necessary to install and load the required R packages. These packages cover different tasks in spatial data handling, attribute management, visualization, and clustering.

**Packages used in this exercise:**\
- **Spatial data handling:** `sf`, `rgdal`, `spdep`\
- **Attribute data handling:** `tidyverse` (includes `readr`, `ggplot2`, `dplyr`, and others such as `tidyr`)\
- **Choropleth mapping:** `tmap`\
- **Multivariate visualization and analysis:** `corrplot`, `ggpubr`, `heatmaply`\
- **Cluster analysis:** `cluster`, `ClustGeo`

```{r}
pacman::p_load(spdep, tmap, sf, ClustGeo, 
               ggpubr, cluster, factoextra, NbClust,
               heatmaply, corrplot, psych, tidyverse, GGally)
```

### 12.4 Data Import and Preparation

#### 12.4.1 Importing Geospatial Data

In this step, the **Myanmar Township Boundary GIS data** (in ESRI Shapefile format) is imported into the R environment using the `st_read()` function from the **sf** package.

```{r}
shan_sf <- st_read(dsn = "data/geospatial", 
                   layer = "myanmar_township_boundaries") %>%
  filter(ST %in% c("Shan (East)", "Shan (North)", "Shan (South)")) %>%
  select(c(2:7))
```

The imported object is named **`shan_sf`**, and it is stored in a **simple feature data frame**.\
We can check the contents of this new object by calling it directly.

```{r}
shan_sf
```

Notice that sf.data.frame is conformed to Hardy Wickham’s [tidy](https://edzer.github.io/rstudio_conf/#1) framework.

Since *shan_sf* is conformed to tidy framework, we can also *glimpse()* to reveal the data type of it’s fields.

```{r}
glimpse(shan_sf)
```

### **12.4.2 Importing aspatial data into R environment**

The csv file will be import using *read_csv* function of **readr** package.

The code chunks used are shown below:

```{r}
ict <- read_csv ("data/aspatial/Shan-ICT.csv")
```

The imported InfoComm variables are extracted from **The 2014 Myanmar Population and Housing Census Myanmar**. The attribute data set is called *ict*. It is saved in R’s \* tibble data.frame\* format.

The code chunk below reveal the summary statistics of *ict* data.frame.

```{r}
summary(ict)
```

There are a total of eleven fields and 55 observation in the tibble data.frame.

### **12.4.3 Derive new variables using dplyr package**

The unit of measurement of the values are number of household. Using these values directly will be bias by the underlying total number of households. In general, the townships with relatively higher total number of households will also have higher number of households owning radio, TV, etc.

In order to overcome this problem, we will derive the penetration rate of each ICT variable by using the code chunk below.

```{r}
ict_derived <- ict %>%
  mutate(`RADIO_PR` = `Radio`/`Total households`*1000) %>%
  mutate(`TV_PR` = `Television`/`Total households`*1000) %>%
  mutate(`LLPHONE_PR` = `Land line phone`/`Total households`*1000) %>%
  mutate(`MPHONE_PR` = `Mobile phone`/`Total households`*1000) %>%
  mutate(`COMPUTER_PR` = `Computer`/`Total households`*1000) %>%
  mutate(`INTERNET_PR` = `Internet at home`/`Total households`*1000) %>%
  rename(`DT_PCODE` =`District Pcode`,`DT`=`District Name`,
         `TS_PCODE`=`Township Pcode`, `TS`=`Township Name`,
         `TT_HOUSEHOLDS`=`Total households`,
         `RADIO`=`Radio`, `TV`=`Television`, 
         `LLPHONE`=`Land line phone`, `MPHONE`=`Mobile phone`,
         `COMPUTER`=`Computer`, `INTERNET`=`Internet at home`) 
```

Let us review the summary statistics of the newly derived penetration rates using the code chunk below.

```{r}
summary(ict_derived)
```

## 12.5 Exploratory Data Analysis (EDA)

### 12.5.1 EDA using Statistical Graphics

After joining the ICT indicators, six new fields were added to the dataset:

-   `RADIO_PR`

-   `TV_PR`

-   `LLPHONE_PR`

-   `MPHONE_PR`

-   `COMPUTER_PR`

-   `INTERNET_PR`

As a first step in exploring these variables, statistical graphics can be used to examine their distributions.\
For example, a **histogram** helps identify whether the values are normally distributed, left-skewed, or right-skewed.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

Boxplot is useful to detect if there are outliers.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

Next, we will also plotting the distribution of the newly derived variables (i.e. Radio penetration rate) by using the code chunk below.

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

```{r}
ggplot(data=ict_derived, 
       aes(x=`RADIO_PR`)) +
  geom_boxplot(color="black", 
               fill="light blue")
```

What can you observed from the distributions reveal in the histogram and boxplot.

The histograms and boxplots reveal **heterogeneity and skewness** in ICT ownership, with modern technologies (computer, internet) concentrated in a few areas, while older technologies (radio, TV) are more evenly distributed.

The code chunks below are used to create the data visualisation. We will create the individual histograms using the code chunk below.

```{r}
radio <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

tv <- ggplot(data=ict_derived, 
             aes(x= `TV_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

llphone <- ggplot(data=ict_derived, 
             aes(x= `LLPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

mphone <- ggplot(data=ict_derived, 
             aes(x= `MPHONE_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

computer <- ggplot(data=ict_derived, 
             aes(x= `COMPUTER_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

internet <- ggplot(data=ict_derived, 
             aes(x= `INTERNET_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

💡 Use the [*ggarrange()*](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html) function of [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/index.html) package is used to group these histograms together.

In the figure below, multiple histograms are plotted to reveal the distribution of the selected variables in the *ict_derived* data.frame.

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet, 
          ncol = 3, 
          nrow = 2)
```

### 12.5.2 EDA using Choropleth Map

#### 12.5.2.1 Joining Geospatial and Aspatial Data

Before preparing a choropleth map, the **geospatial data object** (`shan_sf`) needs to be combined with the **aspatial attribute data** (`ict_derived`).

This is done using the [`left_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html) function from the **dplyr** package.\
- The **base dataset** is the `shan_sf` simple feature data frame (contains township geometries).\
- The **join table** is `ict_derived` (contains ICT attributes).\
- The **unique identifier** used for merging is `TS_PCODE`.

```{r}
shan_sf <- left_join(shan_sf, 
                     ict_derived, by=c("TS_PCODE"="TS_PCODE"))
  
```

```{r}
#| eval: false
write_rds(shan_sf, "data/rds/shan_sf.rds")
```

#### 12.5.2.2 Preparing a Choropleth Map

After the join, the ICT variables from `ict_derived` are now part of the `shan_sf` object.\
It is important to note that **no new dataset is created**; instead, the attributes are updated within the existing `shan_sf` simple feature data frame.

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

#### 12.5.2.2 Preparing a choropleth map

As a quick exploration, a **choropleth map** can be drawn to visualize the distribution of the **Radio penetration rate (RADIO_PR)** across Shan State townships. Using the `qtm()` function from the **tmap** package.

```{r}
qtm(shan_sf, "RADIO_PR")
```

In order to reveal the distribution shown in the choropleth map above are bias to the underlying total number of households at the townships, we will create two choropleth maps, one for the total number of households (i.e. TT_HOUSEHOLDS.map) and one for the total number of household with Radio (RADIO.map) by using the code chunk below.

```{r}
TT_HOUSEHOLDS.map <- tm_shape(shan_sf) + 
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks", 
          title = "Total households") + 
  tm_borders(alpha = 0.5) 

RADIO.map <- tm_shape(shan_sf) + 
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio ") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp=NA, ncol=2)
```

Notice that the choropleth maps above clearly show that townships with relatively larger number ot households are also showing relatively higher number of radio ownership.

Now let us plot the choropleth maps showing the dsitribution of total number of households and Radio penetration rate by using the code chunk below.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style="jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom"))+
  tm_layout(outer.margins=0, asp=0)
```

**Can you identify the differences?**

The two choropleth maps highlight important **differences**:

-   Total households (left map): Darker shades are concentrated in areas with larger populations, meaning radio ownership numbers may appear higher simply because more households exist.

-   Radio penetration rate (right map): This map adjusts for population size, showing the proportion of households owning radios. Some townships with many households no longer appear as dark, while smaller-population townships with relatively high radio access stand out.

## **12.6 Correlation Analysis**

Before we perform cluster analysis, it is important for us to ensure that the cluster variables are not highly correlated.

In this section, you will learn how to use [*corrplot.mixed()*](https://cran.r-project.org/web/packages/corrplot/corrplot.pdf) function of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html)package to visualise and analyse the correlation of the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse", 
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

The correlation plot above shows that COMPUTER_PR and INTERNET_PR are highly correlated. This suggest that only one of them should be used in the cluster analysis instead of both.

## **12.7 Hierarchy Cluster Analysis**

In this section, you will learn how to perform hierarchical cluster analysis. The analysis consists of four major steps:

### **12.7.1 Extracting clustering variables**

The code chunk below will be used to extract the clustering variables from the *shan_sf* simple feature object into data.frame.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars,10)
```

Notice that the final clustering variables list does not include variable INTERNET_PR because it is highly correlated with variable COMPUTER_PR.

Next, we need to change the rows by township name instead of row number by using the code chunk below

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars,10)
```

Notice that the row number has been replaced into the township name.

Now, we will delete the TS.x field by using the code chunk below.

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

```{r}
#| eval: false
write_rds(shan_ict, "data/rds/shan_ict.rds")
```

### 12.7.2 Data Standardisation

In cluster analysis, multiple variables are usually included. Since these variables may have very different value ranges, the clustering results could be biased towards variables with larger scales.\
To reduce this effect, it is important to **standardise the input variables** before running the analysis.

### 12.7.3 Min-Max Standardisation

One common method is **Min-Max standardisation**, which rescales each variable into the range \[0,1\].\
This ensures that all variables contribute equally to the clustering process.

In this exercise, the `normalize()` function from the **heatmaply** package is used to apply Min-Max transformation to the selected clustering variables. The `summary()` function is then applied to check the descriptive statistics of the transformed data.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)

```

Notice that the values range of the Min-max standardised clustering variables are 0-1 now.

### **12.7.4 Z-score standardisation**

Z-score standardisation can be performed easily by using [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of Base R. The code chunk below will be used to stadardisation the clustering variables by using Z-score method.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)

```

Notice the mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.

**Note:** [*describe()*](https://www.rdocumentation.org/packages/Hmisc/versions/4.4-0/topics/describe) of [**psych**](https://cran.r-project.org/web/packages/psych/index.html) package is used here instead of *summary()* of Base R because the earlier provides standard deviation.

***Warning: Z-score standardisation method should only be used if we would assume all variables come from some normal distribution.***

### **12.7.5 Visualising the standardised clustering variables**

Beside reviewing the summary statistics of the standardised clustering variables, it is also a good practice to visualise their distribution graphical.

The code chunk below plot the scaled *Radio_PR* field.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

What statistical conclusion can you draw from the histograms above?

The three histograms show the distribution of **RADIO_PR** under different standardisation methods. The raw values have a wide range (0–500) with right skewness, which may bias clustering. Min–Max standardisation rescales the data to 0–1 while preserving the distribution shape. Z-score standardisation centers the data at mean 0 with unit variance, spreading most values within -2 to +2. Both methods remove scale effects, making variables comparable for clustering.

```{r}
r <- ggplot(data=ict_derived, 
             aes(x= `RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data=shan_ict_s_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data=shan_ict_z_df, 
       aes(x=`RADIO_PR`)) +
  geom_density(color="black",
               fill="light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### **12.7.6 Computing proximity matrix**

In R, many packages provide functions to calculate distance matrix. We will compute the proximity matrix by using [*dist()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of R.

*dist()* supports six distance proximity calculations, they are: **euclidean, maximum, manhattan, canberra, binary and minkowski**. The default is *euclidean* proximity matrix.

The code chunk below is used to compute the proximity matrix using *euclidean* method.

```{r}
#| warning: false
proxmat <- dist(shan_ict, method = 'euclidean')
```

The code chunk below can then be used to list the content of *proxmat* for visual inspection.

```{r}
proxmat
```

### 12.7.7 Computing hierarchical clustering

In R, several packages provide hierarchical clustering functions.\
In this hands-on exercise, the **hclust()** function from the `stats` package is used.

The **hclust()** function applies an *agglomerative* method, where clusters are formed step by step by joining the closest pairs of observations.\
It supports eight different linkage algorithms:

-   `ward.D`\
-   `ward.D2`\
-   `single`\
-   `complete`\
-   `average` (UPGMA)\
-   `mcquitty` (WPGMA)\
-   `median` (WPGMC)\
-   `centroid` (UPGMC)

In this exercise, we will use the **ward.D** method. The result is stored as an object of class **hclust**, which describes the clustering tree (dendrogram) generated in the process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

We can then plot the tree by using *plot()* of R Graphics as shown in the code chunk below.

```{r}
plot(hclust_ward, cex = 0.6)
```

### **12.7.8 Selecting the optimal clustering algorithm**

One of the challenge in performing hierarchical clustering is to identify stronger clustering structures. The issue can be solved by using use [*agnes()*](https://www.rdocumentation.org/packages/cluster/versions/2.1.0/topics/agnes)function of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like *hclus()*, however, with the *agnes()* function you can also get the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

The code chunk below will be used to compute the agglomerative coefficients of all hierarchical clustering algorithms.

```{r}
m <- c( "average", "single", "complete", "ward")
names(m) <- c( "average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output above, we can see that Ward’s method provides the strongest clustering structure among the four methods assessed. Hence, in the subsequent analysis, only Ward’s method will be used.

### **12.7.9 Determining Optimal Clusters**

Another technical challenge face by data analyst in performing clustering analysis is to determine the optimal clusters to retain.

There are [three](https://statweb.stanford.edu/~gwalther/gap) commonly used methods to determine the optimal clusters, they are:

-   [Elbow Method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))

-   [Average Silhouette Method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)

-   [Gap Statistic Method](http://www.web.stanford.edu/~hastie/Papers/gap.pdf)

#### 12.7.9.1 Gap Statistic Method

The **gap statistic** is used to determine the optimal number of clusters. It works by comparing the **within-cluster variation** for different values of *k* with their expected values under a null reference distribution (i.e., data with no obvious clustering). The optimal number of clusters is the value of *k* that **maximizes the gap statistic**, indicating that the observed clustering structure is far from what would be expected by chance.

In this exercise, the function **clusGap()** from the `cluster` package will be used to compute the gap statistic and identify the most suitable number of clusters.

```{r}
#| warning: false
set.seed(12345)

gap_stat <- clusGap(
  shan_ict.std,        
  FUN = hcut,        
  nstart = 25,
  K.max = 10,
  B = 50
)

print(gap_stat, method = "firstmax")
```

Also note that the [*hcut*](https://rpkgs.datanovia.com/factoextra/reference/hcut.html) function used is from [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.

Next, we can visualise the plot by using [*fviz_gap_stat()*](https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html) of [**factoextra**](https://rpkgs.datanovia.com/factoextra/index.html)package.

```{r}
fviz_gap_stat(gap_stat)
```

With reference to the gap statistic graph, the recommended number of clusters is **1**.\
However, retaining only one cluster is not meaningful in practice because clustering is intended to partition the data into multiple groups.

By closely examining the gap statistic plot, we find that the **6-cluster solution** provides the largest gap statistic after the trivial case of one cluster.\
Thus, the **6-cluster solution** is considered the next best choice and is selected for further analysis.

*Note:* In addition to the gap statistic method, the **NbClust** package (Charrad et al., 2014) provides up to **30 different indices** for determining the optimal number of clusters.\
It can recommend the best clustering scheme by testing different combinations of cluster numbers, distance measures, and clustering methods.

### 12.7.10 Interpreting the dendrograms

In the dendrogram, **each leaf represents a single observation**. As we move upward along the tree, observations that are similar to each other are merged into branches. These branches are then fused at higher levels, forming larger clusters. The **height of the fusion** on the vertical axis indicates the level of (dis)similarity between the merged observations or clusters.

-   A **lower height** means the observations are more similar.

-   A **higher height** indicates they are less similar.

It is important to note that conclusions about the proximity of two observations can be drawn only from the **vertical height** where their branches first meet.The **horizontal distance** between two leaves should not be interpreted as similarity.

Additionally, the dendrogram can be enhanced by drawing borders around the selected clusters using the **rect.hclust()** function from the `stats` package.\
The argument `border` allows you to specify the rectangle colors for each cluster.

```{r}
plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward, 
            k = 6, 
            border = 2:5)
```

### **12.7.11 Visually-driven hierarchical clustering analysis**

In this section, we will learn how to perform visually-driven hiearchical clustering analysis by using [*heatmaply*](https://cran.r-project.org/web/packages/heatmaply/index.html) package.

With **heatmaply**, we are able to build both highly interactive cluster heatmap or static cluster heatmap.

#### 12.7.11.1 Transforming the data frame into a matrix

The data was loaded into a data frame, but it has to be a data matrix to make your heatmap.

The code chunk below will be used to transform *shan_ict* data frame into a data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### 12.7.11.2 Plotting interactive cluster heatmap using *heatmaply()*

In the code chunk below, the [*heatmaply()*](https://talgalili.github.io/heatmaply/reference/heatmaply.html) of [heatmaply](https://talgalili.github.io/heatmaply/index.html) package is used to build an interactive cluster heatmap.

```{r}
#| warning: false
heatmaply(normalize(shan_ict_mat),
          Colv=NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA,200,60,NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main="Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### **12.7.12 Mapping the clusters formed**

With closed examination of the dendragram above, we have decided to retain six clusters.

[*cutree()*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R Base will be used in the code chunk below to derive a 6-cluster model.

```{r}
groups <- as.factor(cutree(hclust_ward, k=6))
```

The output is called *groups*. It is a *list* object.

In order to visualise the clusters, the *groups* object need to be appended onto *shan_sf* simple feature object.

The code chunk below form the join in three steps:

-   the *groups* list object will be converted into a matrix;

-   *cbind()* is used to append *groups* matrix onto shan_sf to produce an output simple feature object called `shan_sf_cluster`; and

-   *rename* of **dplyr** package is used to rename *as.matrix.groups* field as *CLUSTER*.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER`=`as.matrix.groups.`)
```

Next, *qtm()* of **tmap** package is used to plot the choropleth map showing the cluster formed.

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

The choropleth map above reveals the clusters are very fragmented. The is one of the major limitation when non-spatial clustering algorithm such as hierarchical cluster analysis method is used.

## **12.8 Spatially Constrained Clustering: SKATER approach**

In this section, you will learn how to derive spatially constrained cluster by using [*skater()*](https://r-spatial.github.io/spdep/reference/skater.html) method of [**spdep**](https://r-spatial.github.io/spdep/index.html) package.

### **12.8.1 Converting into SpatialPolygonsDataFrame**

First, we need to convert `shan_sf` into SpatialPolygonsDataFrame. This is because SKATER function only support **sp** objects such as SpatialPolygonDataFrame.

The code chunk below uses [*as_Spatial()*](https://r-spatial.github.io/sf/reference/coerce-methods.html) of **sf** package to convert *shan_sf* into a SpatialPolygonDataFrame called *shan_sp*.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### **12.8.2 Computing Neighbour List**

Next, [poly2nd()](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package will be used to compute the neighbours list from polygon list.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

We can plot the neighbours list on shan_sp by using the code chunk below. Since we now can plot the community area boundaries as well, we plot this graph on top of the map. The first plot command gives the boundaries. This is followed by the plot of the neighbor list object, with coordinates applied to the original SpatialPolygonDataFrame (Shan state township boundaries) to extract the centroids of the polygons. These are used as the nodes for the graph representation. We also set the color to blue and specify add=TRUE to plot the network on top of the boundaries.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
plot(st_geometry(shan_sf), 
     border=grey(.5))
plot(shan.nb,
     coords, 
     col="blue", 
     add=TRUE)
```

Note that if you plot the network first and then the boundaries, some of the areas will be clipped. This is because the plotting area is determined by the characteristics of the first plot. In this example, because the boundary map extends further than the graph, we plot it first.

### **12.8.3 Computing minimum spanning tree**

#### 12.8.3.1 Calculating edge costs

Next, [*nbcosts()*](https://r-spatial.github.io/spdep/reference/nbcosts.html) of **spdep** package is used to compute the cost of each edge. It is the distance between it nodes. This function compute this distance using a data.frame with observations vector in each node.

The code chunk below is used to compute the cost of each edge.

```{r}
#| warning: false
lcosts <- nbcosts(shan.nb, shan_ict)
```

For each observation, this gives the pairwise dissimilarity between its values on the five variables and the values for the neighbouring observation (from the neighbour list). Basically, this is the notion of a generalised weight for a spatial weights matrix.

Next, We will incorporate these costs into a weights object in the same way as we did in the calculation of inverse of distance weights. In other words, we convert the neighbour list to a list weights object by specifying the just computed ***lcosts*** as the weights.

In order to achieve this, [*nb2listw()*](https://r-spatial.github.io/spdep/reference/nb2listw.html) of **spdep** package is used as shown in the code chunk below.

Note that we specify the *style* as **B** to make sure the cost values are not row-standardised.

```{r}
shan.w <- nb2listw(shan.nb, 
                   lcosts, 
                   style="B")
summary(shan.w)
```

### **12.8.4 Computing minimum spanning tree**

The minimum spanning tree is computed by mean of the [*mstree()*](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep**package as shown in the code chunk below.

```{r}
shan.mst <- mstree(shan.w)
```

After computing the MST, we can check its class and dimension by using the code chunk below.

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

Note that the dimension is 54 and not 55. This is because the minimum spanning tree consists on n-1 edges (links) in order to traverse all the nodes.

We can display the content of *shan.mst* by using *head()* as shown in the code chunk below.

```{r}
head(shan.mst)
```

The plot method for the MST include a way to show the observation numbers of the nodes in addition to the edge. As before, we plot this together with the township boundaries. We can see how the initial neighbour list is simplified to just one edge connecting each of the nodes, while passing through all the nodes.

```{r}
#| height: 9
#| width: 9
plot(st_geometry(shan_sf), 
                 border=gray(.5))
plot.mst(shan.mst, 
         coords, 
         col="blue", 
         cex.lab=0.7, 
         cex.circles=0.005, 
         add=TRUE)
```

### **12.8.5 Computing spatially constrained clusters using SKATER method**

The code chunk below compute the spatially constrained cluster using [*skater()*](https://r-spatial.github.io/spdep/reference/skater.html) of **spdep** package.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2], 
                 data = shan_ict, 
                 method = "euclidean", 
                 ncuts = 5)
```

```{r}
table(clust6$groups)

```

The *skater()* takes three mandatory arguments: - the first two columns of the MST matrix (i.e. not the cost), - the data matrix (to update the costs as units are being grouped), and - the number of cuts. Note: It is set to **one less than the number of clusters**. So, the value specified is **not** the number of clusters, but the number of cuts in the graph, one less than the number of clusters.

The result of the *skater()* is an object of class **skater**. We can examine its contents by using the code chunk below.

```{r}
str(clust6)
```

The most interesting component of this list structure is the groups vector containing the labels of the cluster to which each observation belongs (as before, the label itself is arbitary). This is followed by a detailed summary for each of the clusters in the edges.groups list. Sum of squares measures are given as ssto for the total and ssw to show the effect of each of the cuts on the overall criterion.

We can check the cluster assignment by using the conde chunk below.

```{r}
ccs6 <- clust6$groups
ccs6
```

We can find out how many observations are in each cluster by means of the table command. Parenthetially, we can also find this as the dimension of each vector in the lists contained in edges.groups. For example, the first list has node with dimension 12, which is also the number of observations in the first cluster.

```{r}
table(ccs6)
```

Lastly, we can also plot the pruned tree that shows the five clusters on top of the townshop area.

```{r}
plot(st_geometry(shan_sf), 
     border=gray(.5))
plot(clust6, 
     coords, 
     cex.lab=.7,
     groups.colors=c("red","green","blue", "brown", "pink"),
     cex.circles=0.005, 
     add=TRUE)
```

### **12.8.6 Visualising the clusters in choropleth map**

The code chunk below is used to plot the newly derived clusters by using SKATER method.

```{r}
groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER`=`as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

For easy comparison, it will be better to place both the hierarchical clustering and spatially constrained hierarchical clustering maps next to each other.

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") + 
  tm_borders(alpha = 0.5) 

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") + 
  tm_borders(alpha = 0.5) 

tmap_arrange(hclust.map, shclust.map,
             asp=NA, ncol=2)
```

## **12.9 Spatially Constrained Clustering: ClustGeo Method**

In this section, you will gain hands-on experience on using functions provided by **ClustGeo** package to perform non-spatially constrained hierarchical cluster analysis and spatially constrained cluster analysis.

### **12.9.1 A short note about ClustGeo package**

[**ClustGeo**](https://cran.r-project.org/web/packages/ClustGeo/) package is an R package specially designed to support the need of performing spatially constrained cluster analysis. More specifically, it provides a Ward-like hierarchical clustering algorithm called `hclustgeo()` including spatial/geographical constraints.

In the nutshell, the algorithm uses two dissimilarity matrices D0 and D1 along with a mixing parameter alpha, whereby the value of alpha must be a real number between \[0, 1\]. D0 can be non-Euclidean and the weights of the observations can be non-uniform. It gives the dissimilarities in the **attribute/clustering variable space**. D1, on the other hand, gives the dissimilarities in the **constraint space**. The criterion minimised at each stage is a convex combination of the homogeneity criterion calculated with D0 and the homogeneity criterion calculated with D1.

The idea is then to determine a value of alpha which increases the spatial contiguity without deteriorating too much the quality of the solution based on the variables of interest. This need is supported by a function called `choicealpha()`.

### **12.9.2 Ward-like hierarchical clustering: ClustGeo**

ClustGeo package provides function called `hclustgeo()` to perform a typical Ward-like hierarchical clustering just like `hclust()` you learned in previous section.

To perform non-spatially constrained hierarchical clustering, we only need to provide the function a dissimilarity matrix as shown in the code chunk below.

```{r}
nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster, 
            k = 6, 
            border = 2:5)
```

Note that the dissimilarity matrix must be an object of class `dist`, i.e. an object obtained with the function `dist()`. For sample code chunk, please refer to [5.7.6 Computing proximity matrix](https://r4gdsa.netlify.app/chap05#computing-proximity-matrix)

#### 12.9.2.1 Mapping the clusters formed

Similarly, we can plot the clusters on a categorical area shaded map by using the steps we learned in 5.7.12 Mapping the clusters formed.

```{r}
groups <- as.factor(cutree(nongeo_cluster, k=6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### **12.9.3 Spatially Constrained Hierarchical Clustering**

Before we can performed spatially constrained hierarchical clustering, a spatial distance matrix will be derived by using [`st_distance()`](https://r-spatial.github.io/sf/reference/geos_measures.html) of sf package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

Notice that `as.dist()` is used to convert the data frame into matrix.

Next, `choicealpha()` will be used to determine a suitable value for the mixing parameter alpha as shown in the code chunk below.

```{r}
cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K=6, graph = TRUE)
```

With reference to the graphs above, alpha = 0.2 will be used as shown in the code chunk below.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

Next, `cutree()` is used to derive the cluster objecct.

```{r}
groups <- as.factor(cutree(clustG, k=6))
```

We will then join back the group list with *shan_sf* polygon feature data frame by using the code chunk below.

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

We can now plot the map of the newly delineated spatially constrained clusters.

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## **12.10 Visual Interpretation of Clusters**

### **12.10.1 Visualising individual clustering variable**

Code chunk below is used to reveal the distribution of a clustering variable (i.e RADIO_PR) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

The boxplot reveals Cluster 3 displays the highest mean Radio Ownership Per Thousand Household. This is followed by Cluster 2, 1, 4, 6 and 5.

### **12.10.2 Multivariate Visualisation**

Past studies shown that parallel coordinate plot can be used to reveal clustering variables by cluster very effectively. In the code chunk below, [`ggparcoord()`](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [**GGally**](https://ggobi.github.io/ggally/index.html) package

```{r}
ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

The parallel coordinate plot above reveals that households in Cluster 4 townships tend to own the highest number of TV and mobile-phone. On the other hand, households in Cluster 5 tends to own the lowest of all the five ICT.

Note that the `scale` argument of `ggparcoor()` provide several methods to scale the clustering variables. They are:

-   std: univariately, subtract mean and divide by standard deviation.

-   robust: univariately, subtract median and divide by median absolute deviation.

-   uniminmax: univariately, scale so the minimum of the variable is zero, and the maximum is one.

-   globalminmax: no scaling is done; the range of the graphs is defined by the global minimum and the global maximum.

-   center: use uniminmax to standardize vertical height, then center each variable at a value specified by the scaleSummary param.

-   centerObs: use uniminmax to standardize vertical height, then center each variable at the value of the observation specified by the centerObsID param

There is no one best scaling method to use. You should explore them and select the one that best meet your analysis need.

Last but not least, we can also compute the summary statistics such as mean, median, sd, etc to complement the visual interpretation.

In the code chunk below, `group_by()` and `summarise()` of dplyr are used to derive mean values of the clustering variables.

```{r}
shan_sf_ngeo_cluster %>% 
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```
