---
title: "Hands-on Exercise 05"
---

## 5 Second-Order Spatial Point Pattern Analysis Methods

### 5.1 Overview

Second-order spatial point pattern analysis focuses on the spatial relationships between points, asking whether the presence of one point influences the location of others.\
Unlike first-order analysis, which only looks at point density, second-order methods examine clustering, dispersion, or randomness across different spatial scales.

In this hands-on exercise, we will use *spatstat* functions to explore the spatial processes of childcare centres in Singapore.

The main questions are:\
1. Are childcare centres randomly distributed across Singapore?\
2. If not, where are the areas with higher concentrations of centres?

### 5.2 The Data

To answer these questions, two datasets are used:

-   **Child Care Services (data.gov.sg)** – A point dataset providing both location and attribute information of childcare centres.
-   **Master Plan 2019 Subzone Boundary (URA)** – A polygon dataset describing Singapore’s planning subzone boundaries.

Both datasets are available from Singapore’s open data portal in **KML** and **GeoJSON** formats.

### 5.3 Installing and Loading the R Packages

For this exercise, several R packages are required:

-   **sf** – for importing, managing, and processing vector-based geospatial data.\
-   **spatstat** – provides a wide range of functions for point pattern analysis, including 1st- and 2nd-order methods as well as kernel density estimation (KDE).\
-   **tmap** – for creating cartographic-quality static maps and interactive maps (via the Leaflet API).\
-   **tidyverse** – a family of packages for modern data science, offering consistent and intuitive tools for data manipulation, analysis, and visualization.

The code below installs (if necessary) and loads these packages:

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse,rvest)
```

## **5.4 Data Import and Preparation**

### **5.4.1 Data Import**

The processes of importing and preparing the geospatial data to meet the analysis are similar to Chapter 4. Please refer to the following sub-sections, if necessary:

-   [Importing and Wrangling Geospatial Data Sets](file:///C:/tskam/r4gdsa/_book/chap04.html#importing-and-wrangling-geospatial-data-sets)

```{r}
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaKML.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>% st_transform(crs = 3414)
```

```{r}
childcare_sf <- st_read("data/ChildCareServices.kml") %>% 
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

### **5.4.2 Data Preparation**

Please refer to the following sub-sections, if necessary:

-   [Geospatial Data wrangling](file:///C:/tskam/r4gdsa/_book/chap04.html#geospatial-data-wrangling) for preparing geospatial data in ppp object class required by spatstat package.

```{r}
extract_kml_field <- function(html_text, field_name) {
  if (is.na(html_text) || html_text == "") return(NA_character_)
  
  page <- read_html(html_text)
  rows <- page %>% html_elements("tr")
  
  value <- rows %>%
    keep(~ html_text2(html_element(.x, "th")) == field_name) %>%
    html_element("td") %>%
    html_text2()
  
  if (length(value) == 0) NA_character_ else value
}
```

```{r}
mpsz_sf <- mpsz_sf %>%
  mutate(
    REGION_N = map_chr(Description, extract_kml_field, "REGION_N"),
    PLN_AREA_N = map_chr(Description, extract_kml_field, "PLN_AREA_N"),
    SUBZONE_N = map_chr(Description, extract_kml_field, "SUBZONE_N"),
    SUBZONE_C = map_chr(Description, extract_kml_field, "SUBZONE_C")
  ) %>%
  select(-Name, -Description) %>%
  relocate(geometry, .after = last_col())
```

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

```{r}
write_rds(mpsz_cl, 
          "data/mpsz_cl.rds")
```

#### **5.4.2.1Data wrangling**

-   **Converting sf data frames to ppp class**

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

```{r}
class(childcare_ppp)
```

```{r}
summary(childcare_ppp)
```

-   **Creating *owin* object**

```{r}
sg_owin <- as.owin(mpsz_cl)
```

```{r}
class(sg_owin)
```

-   **Combining point events object and owin object**

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

```{r}
childcareSG_ppp
```

-   **Choa Chu Kang（**childcare_ck_ppp）

```{r}
ck_sf <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
ck_owin <- as.owin(ck_sf)
childcare_ck_ppp <- childcareSG_ppp[ck_owin]
```

-   **Tampines**（childcare_tm_ppp）

```{r}
tm_sf <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
tm_owin <- as.owin(tm_sf)
childcare_tm_ppp <- childcareSG_ppp[tm_owin]
```

## **5.5 Second-order Spatial Point Patterns Analysis**

## **5.6 Analysing Spatial Point Process Using G-Function**

The G function measures the distribution of the distances from an arbitrary event to its nearest event. In this section, you will learn how to compute G-function estimation by using [*Gest()*](https://rdrr.io/cran/spatstat/man/Gest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using [*envelope()*](https://rdrr.io/cran/spatstat/man/envelope.html) of **spatstat**package.

### **5.6.1 Choa Chu Kang planning area**

#### 5.6.1.1 Computing G-function estimation

```{r}
set.seed(1234)
```

The code chunk below is used to compute G-function using *Gest()* of **spatat**package.

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

#### 5.6.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

-   The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

(Monte Carlo test with G-fucntion)

```{r}
# Run once in Console to generate and save
#G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
#saveRDS(G_CK.csr, "data/G_CK_csr.rds")

# In Rmd, always read saved result
G_CK.csr <- readRDS("data/G_CK_csr.rds")
```

```{r}
plot(G_CK.csr)
```

### **5.6.2 Tampines planning area**

#### 5.6.2.1 Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

#### 5.6.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

-   The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
# Run once in Console to generate and save
#G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
#saveRDS(G_tm.csr, "data/G_tm_csr.rds")

# In Rmd
G_tm.csr <- readRDS("data/G_tm_csr.rds")
```

```{r}
plot(G_tm.csr)
```

## **5.7 Analysing Spatial Point Process Using F-Function**

The F function estimates the empty space function F(r) or its hazard rate h(r) from a point pattern in a window of arbitrary shape. In this section, you will learn how to compute F-function estimation by using [*Fest()*](https://rdrr.io/cran/spatstat/man/Fest.html) of **spatstat**package. You will also learn how to perform monta carlo simulation test using [*envelope()*](https://rdrr.io/cran/spatstat/man/envelope.html) of **spatstat** package.

### **5.7.1 Choa Chu Kang planning area**

#### 5.7.1.1 Computing F-function estimation

The code chunk below is used to compute F-function using *Fest()* of **spatat**package.

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

### **5.7.2 Performing Complete Spatial Randomness Test**

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

-   The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

(Monte Carlo test with F-fucntion)

```{r}
# Run once in Console
#F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
#saveRDS(F_CK.csr, "data/F_CK_csr.rds")

F_CK.csr <- readRDS("data/F_CK_csr.rds")
```

```{r}
plot(F_CK.csr)
```

### **5.7.3 Tampines planning area**

#### 5.7.3.1 Computing F-function estimation

(Monte Carlo test with F-fucntion)

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

#### 5.7.3.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

-   The null hypothesis will be rejected is p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
# Run once in Console
#F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
#saveRDS(F_tm.csr, "data/F_tm_csr.rds")

F_tm.csr <- readRDS("data/F_tm_csr.rds")
```

```{r}
plot(F_tm.csr)
```

## **5.8 Analysing Spatial Point Process Using K-Function**

K-function measures the number of events found up to a given distance of any particular event. In this section, you will learn how to compute K-function estimates by using [*Kest()*](https://rdrr.io/cran/spatstat/man/Kest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using *envelope()* of spatstat package.

### **5.8.1 Choa Chu Kang planning area**

#### 5.8.1.1 Computing K-fucntion estimate

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

#### 5.8.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

-   The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
# Run once in Console
#K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
#saveRDS(K_ck.csr, "data/K_ck_csr.rds")

K_ck.csr <- readRDS("data/K_ck_csr.rds")
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### **5.8.2 Tampines planning area**

#### 5.8.2.1 Computing K-fucntion estimation

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### 5.8.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

-   The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
# Run once in Console
#K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
#saveRDS(K_tm.csr, "data/K_tm_csr.rds")

K_tm.csr <- readRDS("data/K_tm_csr.rds")
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## **5.9 Analysing Spatial Point Process Using L-Function**

In this section, you will learn how to compute L-function estimation by using [*Lest()*](https://rdrr.io/cran/spatstat/man/Lest.html) of **spatstat** package. You will also learn how to perform monta carlo simulation test using *envelope()* of spatstat package.

### **5.9.1 Choa Chu Kang planning area**

#### 5.9.1.1 Computing L Fucntion estimation

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

#### 5.9.1.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Choa Chu Kang are randomly distributed.

-   H1= The distribution of childcare services at Choa Chu Kang are not randomly distributed.

-   The null hypothesis will be rejected if p-value if smaller than alpha value of 0.001.

The code chunk below is used to perform the hypothesis testing.

```{r}
# Run once in Console
#L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
#saveRDS(L_ck.csr, "data/L_ck_csr.rds")

L_ck.csr <- readRDS("data/L_ck_csr.rds")
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

### **5.9.2 Tampines planning area**

#### 5.9.2.1 Computing L-fucntion estimate

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

#### 5.9.2.2 Performing Complete Spatial Randomness Test

To confirm the observed spatial patterns above, a hypothesis test will be conducted. The hypothesis and test are as follows:

-   Ho = The distribution of childcare services at Tampines are randomly distributed.

-   H1= The distribution of childcare services at Tampines are not randomly distributed.

-   The null hypothesis will be rejected if p-value is smaller than alpha value of 0.001.

The code chunk below will be used to perform the hypothesis testing.

```{r}
# Run once in Console
#L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
#saveRDS(L_tm.csr, "data/L_tm_csr.rds")

L_tm.csr <- readRDS("data/L_tm_csr.rds")
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```

### 5.9 Conclusion

In this chapter, we moved beyond first-order analysis and explored **second-order spatial point pattern methods** to study childcare centres in Singapore. By applying Ripley’s K-function, L-function, and the pair correlation function, we examined whether centres are randomly distributed, clustered, or more evenly spaced at different spatial scales.

The results indicate that childcare centres are **not randomly distributed**. Instead, they show clear evidence of **clustering**, which reflects real-world planning and demand patterns. These findings highlight areas of concentration as well as regions that may be underserved, offering useful insights for both policymakers and service providers when planning new facilities.

Overall, second-order analysis complements first-order methods by uncovering the **relationships between points**, providing a deeper understanding of spatial processes that cannot be captured by density measures alone.
