---
title: "Hands-on Exercise 10: Local Measures of Spatial Autocorrelation"
author: "BingBing"
date: "September 18, 2025"
date-modified: "last-modified"
---

## 10.1 Overview — Local Measures of Spatial Autocorrelation (LMSA)

LMSA assess how each observation relates to its surrounding neighbors, yielding location-specific scores rather than a single map-wide statistic. They share intuition with global measures and, in some cases, the global metric can be decomposed into local components—e.g., LISA (Local Indicators of Spatial Association). In addition to LISA, Getis–Ord 𝐺 𝑖 G i ​

is introduced as an alternative LMSA that offers complementary hot/cold-spot insights.

By the end of this exercise you will be able to:

1.  import geospatial data with sf;

2.  import CSV data with readr;

3.  perform relational joins with dplyr;

4.  compute LISA (to detect local clusters and outliers) with sfdep;

5.  compute Getis–Ord 𝐺 𝑖 (to detect hot/cold spots) with sfdep;

6.  visualize the outputs with tmap.

## 10.2 Getting Started 10.2.1 The analytical question

In spatial policy, a central goal is equitable spatial distribution of development. We ask: (i) Are development levels evenly distributed? If no, (ii) is there spatial clustering? If yes, (iii) where are the clusters? The case study examines GDP per capita in Hunan Province, PRC.

### 10.2.2 The Study Area and Data

-   Area: Hunan Province at county level.

-   Datasets: (1) county administrative boundary shapefile; (2) Hunan_2012.csv containing selected 2012 development indicators.

### 10.2.3 Setting the Analytical Tools

Ensure the following R packages are installed and loaded:

-   sf (import/handle geospatial data),

-   tidyverse (wrangle attribute data),

-   sfdep (compute spatial weights; global and local spatial autocorrelation),

-   tmap (produce cartographic-quality choropleth maps).

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## **10.3 Getting the Data Into R Environment**

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

### **10.3.1 Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

In the code chunk above, `st_transform()` is used to transform Hunan geospatial data from WGS 84 to UTM zone 50N (i.e. EPSG: 32650).

### **10.3.2 Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R data frame class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### **10.3.3 Performing relational join**

The code chunk below will be used to update the attribute table of *hunan’s* sf data frame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### **10.3.4 Visualising Regional Development Indicator**

In the code chunks below, tmap functions are used:

-   to build two choropleth maps by using equal interval (i.e equal) and quantile (i.e. quantile) classification methods, and

-   to plot both maps next to each other by using `tmap_arrange()`.

```{r}
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in(
                  "left", "bottom"))) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile interval classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## **10.4 Local Indicators of Spatial Association (LISA)**

LISA provides **location-specific** statistics to detect **clusters** (high–high, low–low) and **outliers** (high–low, low–high) in a spatial variable. Here, we use **local Moran’s I** to identify where GDP per capita in Hunan (2012) departs from what pure spatial randomness would produce, highlighting counties whose values are unusually similar to—or different from—their neighbors.

### **10.4.1 Computing Contiguity Spatial Weights**

First build neighborhood structure, then convert it to weights. Use **`sfdep::st_contiguity()`** to create a first-order contiguity neighbors list; by default `queen = TRUE` (shared edge **or** corner qualifies). Next, pass the neighbors list to **`sfdep::st_weights()`** to obtain polygon weights.

Next, [`st_weights()`](https://sfdep.josiahparry.com/reference/st_weights) is used to calculate polygon spatial weights from the *nb* list.

`st_weights()` provides three arguments, they are:

-   nb: A neighbor list object as created by `st_neighbors()`.

-   style: Default “W” for row standardized weights. This value can also be “B”, “C”, “U”, “minmax”, and “S”. B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   allow_zero: If TRUE, assigns zero as lagged value to zone without neighbors.

```{r}
wm_q <- hunan %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(
           nb, style = "W"),
         .before = 1) 
```

To reveal the neighbour list, code chunk below is used.

```{r}
summary(wm_q$nb)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbour.

### 10.4.2 Row-standardised weights

Row-standardising (**style = "W"**) gives each neighbor equal weight $1/k$ (where $k$ is the number of neighbors), so a unit’s spatial lag is the **average** of its neighbors’ values. This is simple and comparable across units, but **edge units** (with fewer neighbors) may have lags based on scant information, which can distort perceived autocorrelation. Simplicity wins here, so we keep `"W"`, noting that alternatives like **`"B"`** (binary, unstandardised) exist.

### What the code chunk teaches

-   **Required input:** `nb2listw()` takes an **`nb`** (neighbors list) object.

-   **Two key args:**

    -   **`style`** – weighting scheme:

        -   `"W"` row-standardised (rows sum to 1);
        -   `"B"` binary (0/1 links, unstandardised);
        -   `"C"` globally standardised (all weights sum to *n*);
        -   `"U"` = `"C"` divided by each unit’s neighbor count (rows sum to 1);
        -   `"minmax"` rescaled to $[0,1]$;
        -   `"S"` Tiefelsdorf variance-stabilising (sums to *n*).

    -   **`zero.policy`** – handling **isolates** (no neighbors). If `TRUE`, inserts zero-length weight vectors so the **spatial lag = 0** for those units—sometimes sensible, sometimes not; use with care.

### **10.4.3 Computing local Moran’s I**

To compute local Moran’s I, the [`local_moran()`](https://sfdep.josiahparry.com/reference/local_moran) function of **sfdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran’s I of *GDPPC2012* at the county level.

```{r}
lisa <- wm_q %>% 
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99),
         .before = 1) %>%
  unnest(local_moran)
```

`local_moran()` function returns a matrix of values whose columns are:

-   Ii: the local Moran’s I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

```{r}
glimpse(lisa)
```

#### 10.4.3.1 Mapping the local Moran’s I

#### 10.4.3.2 Mapping local Moran’s I values

Using choropleth mapping functions of **tmap** package, we can plot the local Moran’s I values by using the code chinks below.

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "ii",
              fill.scale = tm_scale_intervals(
                style = "pretty",
                n = 5,
                values = "brewer.RdBu"),
              fill.legend = tm_legend(
                title = "Local Morans'I",
                position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("Loal Morans'I of GDPPC (Queen's method)")
```

#### 10.4.3.3 Mapping local Moran’s I p-values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran’s I p-values by using functions of **tmap** package.

```{r}
tm_shape(lisa) +
  tm_polygons(fill = "p_ii", 
              fill.scale = tm_scale_intervals(
                breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
                values = "-brewer.Reds"),
              fill.legend = tm_legend(
                title = "p-value",
      position = tm_pos_out())) + 
  tm_borders(fill_alpha = 0.5) +
  tm_title("p-values of Loal Moran's I of GDPPC (Queen's method)")
```
